\chapter*{Abstract}

\addcontentsline{toc}{chapter}{Abstract}

Large language models synthesise code that often \emph{looks} correct under
everyday unit tests yet still mishandles thin specification regions---especially
catch-all residuals and ordered overlaps.
Consider a tier classifier whose \texttt{OTHERWISE} branch must return
\texttt{Review}: an LLM candidate can pass every Reject/Pass sample while
silently promoting the mid-band to \texttt{Pass}.
A release dashboard that tracks mean pass rate alone will ship that defect.

We reframe the goal as \textbf{acceptance safety}: release only when every
specification-derived witness matches \emph{and} a structural screen clears
($\mathrm{Accept}\Leftrightarrow\mathrm{Conf}{=}1\wedge\mathrm{Screen}{=}\mathit{pass}$).
The vehicle is a \textbf{conjunctive release decision framework} realised by
HSP-Agile on SOFL/FSF: first-match SMT witnesses, a typed Semantic Feedback IR
for repair under headroom, and a pluggable screen.
Here \emph{FAR} (False Accept Rate) is the fraction of faulty mutants that a
mode incorrectly releases; \emph{PDR} is the complementary rejection rate.

At comparable mean conformance, conjunctive Accept cuts escaped defects
(E2: M FAR 5.0\% vs.\ B2 8.8\%; PDR 95.0\% vs.\ 91.2\%; bootstrap CIs on the
gaps exclude~0).
When repair still has headroom, structured feedback rescues catch-all failures
that unstructured dumps miss (E6: semantic\_ir 86.9\% vs.\ test\_only 79.1\%,
${+}7.7$\,pp; W/L/T $14/4/102$; 95\% CI $[2.3,\,13.6]$\,pp; $p{=}0.018$;
all 14 Conf.\ wins are \texttt{others} rescues, none are ordering repairs).
At matched budget~$K{=}3$, equal-$K$ Conf.\ places \texttt{M\_eq} at 100.0\%
vs.\ B2 97.5\% (${+}2.5$\,pp; $3/0/117$).
Prefer test-feedback (B2) by default; escalate to full mode~M when release
requires $\mathrm{Accept}{=}1$ or FAR${\le}5\%$ and headroom remains.
Table~\ref{tab:evidence-ladder} summarises the evidence ladder.
Artifacts released.

\paragraph{Keywords.}
Acceptance safety; conjunctive release; specification-guided repair; false
accept rate; LLM synthesis; SOFL; SMT witnesses; catch-all rescue.
