\chapter*{Abstract}

\addcontentsline{toc}{chapter}{Abstract}

Large language models synthesise code that often \emph{looks} correct under
everyday unit tests yet still mishandles thin specification regions---especially
catch-all residuals and ordered overlaps.
Consider a tier classifier whose \texttt{OTHERWISE} branch must return
\texttt{Review}: an LLM candidate can pass every Reject/Pass sample while
silently promoting the mid-band to \texttt{Pass}.
A release dashboard that tracks mean pass rate alone will ship that defect.

We study headroom-conditioned catch-all repair under a SpecIR first-match
oracle, together with a conditional dual-gate release check
($\mathrm{Accept}\Leftrightarrow\mathrm{Conf}{=}1\wedge\mathrm{Screen}{=}\mathit{pass}$).
HSP-Agile implements the check on SOFL/FSF with Priority-Encoded Witness
Generation, Scenario-Indexed Feedback Rendering, and a configurable static
screen.
As a formal-engineering method, the check is a bounded operational criterion
under a faithful SpecIR encoding, not a theorem of program safety or a new
SMT decision procedure.
Here \emph{FAR} (False Accept Rate) is the fraction of faulty mutants that a
mode incorrectly releases; \emph{PDR} is the complementary rejection rate.

When test-only Conf.\ is already~$0$ ($n{=}13$), Semantic IR rescues all tasks
(${+}90.7$\,pp); on the headroom set (Conf.${<}1$, $n{=}114$) the paired mean
is ${+}8.2$\,pp (CI excludes~0).
Among 18 decisive tasks, Semantic IR wins 14/18 (77.8\%).
The all-task ${+}7.7$\,pp figure is only a supporting aggregate.
On industrial-style mutants ($n{=}132$), Screen cuts FAR from 56.8\% to
32.6\%---the primary prevention claim; over the full mutant pool ($n{=}852$)
that lift appears as FAR 8.8\% to 5.0\%, but HardSynthetic mutants are already
saturated and do not exercise Screen.
The all-task ${+}7.7$\,pp aggregate (semantic\_ir 86.9\% vs.\ test\_only 79.1\%;
W/L/T $14/4/102$; 95\% CI $[2.3,\,13.6]$\,pp; $p{=}0.018$) is a supporting
summary, not a universal repair claim, and all 14 Conf.\ wins are
\texttt{others} rescues, none are ordering repairs.
When test-only Conf.\ is already partial ($0{<}\mathrm{Conf}{<}1$), keep B2
rather than forcing Semantic IR (${-}2.5$\,pp in the partial-failure band).
At matched budget~$K{=}3$, equal-$K$ Conf.\ places \texttt{M\_eq} at 100.0\%
vs.\ B2 97.5\% (${+}2.5$\,pp; $3/0/117$).
Gemini hard-seed runs support the feedback mechanism on frozen injected bugs.
Under a frozen external hard-seed protocol (test-only Conf.${<}1$),
\texttt{wrong\_relop} transfers at ${+}20.6$\,pp ($n{=}79$; CI $[12.7,\,28.6]$).
On GitHub harvest reference mutants, Screen cuts FAR from 60.4\% to 11.5\%
(94 Screen-hits); published-industrial screening saturates (0 hits).
The controlled SIFR estimate remains E6 on BENCH-120; external rows are
protocol-transfer diagnostics, not a second headline.
Test-only B2 remains the default.
Escalate when Accept is required or a contractual FAR budget exists \emph{and}
repair headroom remains; Screen helps on industrial-pattern mutants but does
not attain FAR${\le}5\%$ on that slice.
Table~\ref{tab:evidence-ladder} summarises the evidence ladder.
Artifacts released.

\paragraph{Keywords.}
Acceptance checking; dual-gate release; specification-guided repair; false
accept rate; LLM synthesis; SOFL; SMT witnesses; catch-all rescue.
