\section{Background and Related Work}
\label{sec:related}

\subsection{Specification-Conformance Defects in LLM Synthesis}

LLM-assisted code generation has matured rapidly~\cite{chen2021codex,openai2023gpt4},
but empirical studies consistently find that 15--40\% of suggestions contain subtle
semantic defects not caught by coarse test suites~\cite{dakhel2023copilot}.
In the formal specification setting, these defects are qualitatively different from
typical correctness failures: they violate a precisely-stated behavioural contract at
boundary regions that the specification was designed to characterise.
Prior SOFL-oriented studies showed that early fault prevention reduces downstream
correction costs~\cite{li2022qrs,li2023iet}, but they assume human-written
implementations; the integration with LLM-based synthesis creates new failure modes that
neither human-process prevention nor LLM post-processing alone addresses.

\subsection{Verifier-in-the-Loop and CEGIS}

Counterexample-guided inductive synthesis (CEGIS)~\cite{solar2006sketch} established the
canonical template for iterative correction: a verifier supplies counterexamples and a
synthesiser refines its candidate until no counterexample remains.
Verifier-guided LLM synthesis has shown encouraging results in Dafny and related
settings~\cite{lynn2024tosem,dafnybench2024}.
Table~\ref{tab:related-positioning} positions our framework against these approaches on
the dimensions that matter for specification-conformance defect prevention.

\begin{table}[t]
  \caption{Positioning \textsc{SgDP} against related approaches.  A tick indicates
    the property is satisfied; a dash indicates it is not a design goal.}
  \label{tab:related-positioning}
  \small
  \centering
  \begin{tabular}{lcccc}
    \toprule
    & Spec-aware & Structured & Prevention & General \\
    Approach & witnesses & feedback IR & protocol & framework \\
    \midrule
    CEGIS (Solar-Lezama)          & ---  & --- & ---  & \checkmark \\
    DafnyBench~\cite{dafnybench2024}     & \checkmark & --- & --- & --- \\
    Self-Refine~\cite{madaan2023selfrefine}   & --- & --- & --- & \checkmark \\
    Self-Debug~\cite{chen2023selfdebug}    & --- & --- & --- & \checkmark \\
    Reflexion~\cite{shinn2023reflexion}    & --- & --- & --- & \checkmark \\
    CodeT~\cite{chen2022codet}           & --- & --- & --- & --- \\
    \textsc{SgDP} (this paper)    & \checkmark & \checkmark & \checkmark & \checkmark \\
    \bottomrule
  \end{tabular}
\end{table}

The key gap in all prior approaches is the absence of \emph{specification-aware}
witness generation: they use random tests, execution traces, or general symbolic
execution, none of which reliably reaches the low-measure boundary regions where
specification-conformance defects manifest.
Our Constraint-guided Witness Generator closes this gap by encoding the specification's
guard precedence chain as SMT constraints~\cite{barrett2018smt}.

\subsection{Structured Feedback for LLM Repair}

Recent work on LLM repair has explored increasingly rich feedback forms: test
failures~\cite{chen2023selfdebug}, natural-language critique~\cite{madaan2023selfrefine},
execution traces~\cite{shinn2023reflexion}, and compiler diagnostics.
\citet{lynn2024verifierloop} document that test-feedback loops with underspecified
diagnostics converge significantly more slowly on violations requiring structural code
changes.  Our Semantic Feedback IR addresses this by surfacing specification-level
structure — violated scenario identity, guard predicate text, violation type —
that pure test-feedback discards.
Controlled comparison (E6) isolates the contribution of this structure from the
contribution of additional feedback volume.

\subsection{Specification Mutation and Prevention Evaluation}

Specification mutation is a long-established method for evaluating fault sensitivity of
testing and analysis processes~\cite{okun2007nist,circus2017mutation,li2023iet}.
Prior prevention evaluation typically conflates specification-confusion and
implementation-screening failure modes.
Our prevention evaluation protocol separates these into two mutant layers (Section~\ref{sec:method})
and measures both PDR and FAR, providing a more discriminating assessment than simple
pass/fail conformance metrics.

\subsection{Static Analysis for Specification Anti-patterns}

Static analysis tools such as FindBugs~\cite{hovemeyer2004findbugs} target
implementation-level code smells.
Our Pattern Guard component extends this paradigm to the specification-to-code translation
phase: it encodes SOFL-specific requirement anti-patterns (e.g., guard-order confusion,
missing precondition check) as AST and regex analysis passes, producing a
specification-derived defect classifier.
Unlike general bug-finding tools, the patterns are derived directly from the specification
language's fault taxonomy, making their precision and recall measurable against the
mutant population (E7).

\subsection{Formal Methods in Agile Development}

The tension between formal rigour and agile delivery pace is well-studied~\cite{agile2001manifesto}.
Lightweight formal methods — model checking, property-based testing, runtime verification
— have been proposed as practical compromises~\cite{clarke1999modelchecking}.
\textsc{SgDP} contributes to this tradition by demonstrating that specification-grounded
defect prevention can be integrated into a per-sprint CI/CD gate without prohibitive
latency overhead, operating in the space between pure test-feedback and full deductive
verification~\cite{leino2010dafny}.

\subsection{Gap Summary}

Existing approaches either guarantee correctness but require proof annotations that LLMs
systematically fail to produce~\cite{loughman2024dafnybench}, or they repair code without
grounding feedback in specification semantics.  \textsc{SgDP} occupies the practical
middle ground: specification-semantic awareness without proof obligation, with an
evaluation protocol that measures prevention rather than generation accuracy.
