\section{Experimental Setup}
\label{sec:setup}

\subsection{Research Questions}

\textbf{RQ1 (Overall Effectiveness):}
Does specification-guided prevention significantly improve formal correctness over pure
LLM synthesis?
Evaluated via strict formal conformance and PDR/FAR on the 120-task hard benchmark
(E1, E2).

\textbf{RQ2 (Fault Class Prevention):}
What specification defect classes does the framework prevent, and which remain outside
its reach?
Evaluated via per-operator PDR heatmap (E2), per-pattern P/R/F1 (E7), and failure
taxonomy (E9).

\textbf{RQ3 (Component Attribution):}
How much does each semantic component (formal checker, Semantic Feedback IR, pattern
guard) contribute independently?
Evaluated via ablation (A1/A2/A3 vs.\ M, E1), repair dynamics (E5), and feedback
variant comparison (E6).

\textbf{RQ4 (Complexity Scaling):}
How does the framework's advantage scale across specification complexity dimensions?
Evaluated via overlap density tertiles and guard complexity stratification (E3).

\textbf{RQ5 (Quality--Overhead Trade-off):}
What is the quality gain per unit of latency and LLM calls relative to the strongest
baseline?
Evaluated via latency analysis and Pareto comparison (E1).

\subsection{Benchmark Construction}

The benchmark consists of 120 hard synthetic Agile-SOFL tasks, selected from 180
generated candidates by a Z3 satisfiability filter that retains only tasks with
non-empty precedence-sensitive overlap regions (the class where
specification-conformance defects occur).
Each task contains five integer input variables, three output variables, and 8 ordered
FSF scenarios with overlapping guard regions.
Tasks are annotated with three complexity metadata fields: (i)~\emph{overlap density}
(fraction of input space where multiple guards are simultaneously satisfiable);
(ii)~\emph{guard complexity label} (Simple/AND/Nested/Arithmetic/Mixed);
(iii)~\emph{scenario count}.
The template baseline B0 reaches 89.2\% mean formal conformance in one shot
($\approx$0.2\,s per task) on canonical E1---matching B1 at seed~0 and
confirming that strict success (5.0\% under conjunctive Accept) and
prevention metrics, not aggregate mean Conf., discriminate modes on this slice.

An additional 20--30 tasks derived from HumanEval and MBPP problems with multi-case
conditional logic are converted to FSF-style ordered scenarios following a documented
selection and conversion protocol; these are used in the generalisation study (E8).

\subsection{Compared Modes}

\begin{table}[t]
  \caption{Compared modes.  Columns indicate whether each component is active.}
  \label{tab:modes}
  \centering
  \small
  \begin{tabular}{lcccc}
    \toprule
    Mode & Description & Formal checker & Pattern guard & Repair loop \\
    \midrule
    B0 & Reference (template) & eval only & --- & --- \\
    B1 & One-shot LLM & --- & --- & --- \\
    B2 & Test-feedback & --- & --- & \checkmark (tests) \\
    B3 & Self-Refine & --- & --- & \checkmark (self-critique) \\
    B4 & Self-Debug & --- & --- & \checkmark (execution trace) \\
    M  & Full \textsc{SgDP}/HSP-Agile & \checkmark & \checkmark & \checkmark (IR) \\
    A1 & Ablation: no formal checker & --- & \checkmark & \checkmark (IR) \\
    A2 & Ablation: no pattern guard & \checkmark & --- & \checkmark (IR) \\
    A3 & Ablation: no repair loop & \checkmark & \checkmark & --- \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Extended baselines.}
B3 (Self-Refine style~\cite{madaan2023selfrefine}): after each synthesis attempt, the
same LLM is asked to critique its output and revise without any external checker.
B4 (Self-Debug style~\cite{chen2023selfdebug}): the candidate is executed on B0
reference test cases and execution failures are fed back as natural-language without
specification-level annotation.
Both extended baselines use the same prompt template family as M and the same budget $K$.

\subsection{Mutant Design}

Two mutant tracks:
\begin{enumerate}
  \item \textbf{Specification-confusion} (spec layer): SNO (scenario negation/order swap),
    ORO (output reversal), MCO (missing condition), BCO (boundary condition offset).
  \item \textbf{Implementation-screening} (impl layer): ICO (incorrect condition),
    WRO (wrong return order), MBO (missing branch), DRO (dependency reversal).
\end{enumerate}
Each mutant is generated by applying one first-order operator to a known-correct
candidate.  Invalid (non-compilable) mutants are excluded and reported separately.

\subsection{Evaluation Protocol}

\paragraph{E1 — Overall Effectiveness.}
120 tasks $\times$ 9 modes $\times$ $\geq$3 repeats. Primary metrics: Conf, PDR, FAR,
latency. Budget $K=3$, $T_\text{gen}=0.7$, $T_\text{repair}=0.0$.

\paragraph{E2 — Full Prevention Evaluation.}
120 tasks $\times$ 6 modes $\times$ 8 mutant operators, measuring PDR/FAR per operator.

\paragraph{E3 — Specification Complexity.}
Conf stratified by overlap density tertile and guard complexity label.

\paragraph{E4 — Boundary Density.}
Scenario coverage rate of SMT witnesses vs.\ random sampling across 5 case budgets
(4/8/16/32/64) and 3 density tiers.

\paragraph{E5 — Repair Dynamics.}
$\mathrm{Conf}(k)$ trajectory for M vs.\ B2 at $k=1,2,3$ from logged
\texttt{attempt\_history}.

\paragraph{E6 — Feedback Variant Comparison.}
Final Conf and iterations-to-convergence for feedback variants A (test only),
B (test + expected), and C (full Semantic Feedback IR).

\paragraph{E7 — Pattern Guard P/R/F1.}
Per-category precision, recall, and F1 of the pattern guard on the E2 mutant population.

\paragraph{E8 — Generalisation.}
Conf for M, B1, B2 across SOFL/FSF, Mini-Z, and Mini-StateMachine notations.

\paragraph{E9 — Failure Analysis.}
Taxonomy of remaining failures in M mode by category.

\subsection{Statistical Protocol}

\begin{itemize}
  \item Paired Wilcoxon signed-rank test for conformance comparisons.
  \item Mann-Whitney U test for latency comparisons.
  \item Holm-Bonferroni correction applied across all pairwise mode comparisons.
  \item Cliff's $\delta$ as non-parametric effect size.
  \item Bootstrap 95\% confidence intervals ($n=1000$ resamples).
  \item Significance threshold $\alpha = 0.05$.
\end{itemize}

\subsection{Reproducibility}

All results are produced via a scripted pipeline with deterministic seeds.
The \texttt{attempt\_history} field in every result JSONL records per-iteration
conformance and counterexample counts, enabling E5 repair dynamics analysis without
re-running experiments.
Data preparation, aggregation, and plotting are separated into distinct pipeline stages.
Results pending reruns are marked \texttt{TBD}; no estimated or proxy values appear
in reported tables or figures.
