\section{Discussion and Threats to Validity}
\label{sec:threats}

\subsection{Interpretation}
The study is designed to isolate how formal checking, repair feedback, and pattern screening interact under precedence-sensitive scenarios.
Interpretations are limited to the evaluated benchmark and execution protocol; claims about broader industrial impact require additional external studies.

\subsection{Threats to Validity}
\textbf{Construct validity}: Our correctness construct is operationalized through generated concrete scenario cases and strict acceptance rules.
This may not capture all semantically equivalent implementations outside the checked input space.
Mitigation includes explicit scenario ordering, multi-output checks, and complementary prevention metrics (PDR/FAR).\\
\textbf{Internal validity}: LLM outputs are stochastic and may vary across runs.
We mitigate this threat by deterministic seeds, fixed prompt templates, consistent attempt budgets, and script-based execution.
Residual variance is handled through non-parametric statistical tests and effect-size reporting.\\
\textbf{External validity}: The benchmark includes synthetic hard tasks that emphasize precedence-sensitive failure modes.
These tasks improve discrimination but may not represent all industrial domains or specification styles.
Future replications should include additional projects, domains, and team-level workflows.\\
\textbf{Conclusion validity}: Statistical conclusions depend on sample size, run count, and multiple-comparison handling.
We report inferential tests with effect sizes and mark pending quantities as \texttt{TBD} until full reruns are finalized.

\subsection{Implications for Practice}

\textbf{Integration into agile sprint cycles.}
HSP-Agile is designed to be operationally lightweight relative to the time budget of a typical agile sprint.
The per-task pipeline---synthesis, checking, repair, and pattern screening---runs within a bounded attempt budget $K$ and can be parallelized across tasks assigned in a sprint.
In practice, teams can configure $K$ to match delivery pace: a small $K$ gives faster turnaround at the cost of fewer repair attempts, while a larger $K$ trades latency for higher strict success rates.
The pipeline does not require modification of existing sprint tooling; it operates as an acceptance gate between LLM-assisted coding and code review, producing an auditable decision record for each candidate.

\textbf{The formal checker as a low-cost conformance oracle.}
The formal checker in HSP-Agile serves as a specification-grounded oracle that is cheaper to query than a full theorem prover and more semantically precise than ad-hoc unit tests.
By instantiating executable cases directly from FSF scenario guards and expected outputs, it produces concrete counterexamples that are actionable for both the repair loop and the developer.
This oracle role is most valuable when specifications are available in structured form (e.g., SOFL FSF tables): the checker can then flag conformance violations without requiring the developer to manually author test cases.
In settings where specifications are informal or underspecified, the checker's coverage depends on the quality of the case generation heuristic, and supplemental testing remains necessary.

\textbf{Effectiveness of the pattern guard.}
The pattern guard is most effective when failure modes are systematic and encodable as named anti-patterns---for example, branch-order confusion arising from scenario priority misinterpretation, or missing guard refinements in multi-output scenarios.
In the ablation study, removing the pattern guard produces the largest relative drop in prevention detection rate (PDR), confirming its contribution is not redundant with conformance checking alone.
Conversely, the pattern guard is least effective for novel defect classes not yet represented in the pattern catalog, and for specification styles with non-standard guard structures.
Practitioners should treat the pattern set as a living artifact: as new failure modes are observed in production or code review, corresponding patterns can be added incrementally without retraining the LLM or redesigning the checker.
