Ordered guard specifications enforce \emph{first-match} precedence: when
several guards hold, only the earliest scenario may fire.
LLM-generated code often violates that rule on \emph{thin overlap} slices that
random unit tests almost never sample~\cite{ammann2016testing,zhu1997specification}.

Running example \texttt{classify\_signal}: with $\mathit{threshold}{<}0$,
input $(-3,-10)$ satisfies both $G_1$ (\texttt{level < 0}) and $G_2$
(\texttt{level >= threshold}); FSF ordering mandates \texttt{"Error"} via~$S_1$,
yet a typical candidate checks the threshold first and returns
\texttt{"Critical"}.
Pass/fail repair never sees the defect---hence SpecIR witnesses and a typed
Semantic Feedback IR.

\textbf{Main claim.}
On ordered-guard specs, typed Semantic Feedback IR improves repair over
unstructured test-only under \emph{controlled equal budget} (E6, $K{=}3$:
${+}7.7$\,pp; paired 14/4/102; 95\% CI $[2.3,\,13.6]$; $p{=}0.018$);
conjunctive $\mathrm{Accept}$ improves prevention (E2);
\textbf{deploy B2 by default} for mean Conf.\ / latency unless Accept or
PDR/FAR is required (C4).
Equal-$K$ Conf.\ ranking (\texttt{M\_eq} vs.\ B2, both $K{=}3$): ${+}2.5$\,pp
(3/120 wins); strengthened E1 ($K{=}5$ bundle) is a near-ceiling stress-test,
not the lead claim.

\textbf{HSP-Agile} instantiates \textsc{SgDP} for SOFL/FSF with
(i)~SpecIR first-match regions,
(ii)~typed Semantic Feedback IR, and
(iii)~conjunctive $\mathrm{Accept}$ (full witness conformance $\wedge$ clear
high-severity pattern screen).

\textbf{Contributions.}
\textbf{(C1)}~SpecIR is a notation-agnostic hinge for first-match regions
(engineering; E3 tertiles are \emph{non-monotone}---do not use overlap alone).
\textbf{(C2)}~Typed Semantic Feedback IR yields \textbf{+7.7\,pp} over
unstructured test-only under~M (E6; CI excludes~0)---scoped vs.\ length-matched
traces by E14.
\textbf{(C3)}~Conjunctive release improves prevention (E2: M PDR
95.0\%\,/\,FAR 5.0\% vs.\ B2 91.2\%\,/\,8.8\%).
\textbf{(C4)}~Deployment boundary: B2 default on unfiltered/industrial;
escalate to~M iff $\mathrm{Accept}{=}1$ or target FAR${\le}5\%$ is required.
