SOFL/FSF models ordered guards with first-match
precedence~\citep{liu1998sofl,liu2004soflbook}; Agile-SOFL embeds them in
sprint workflows~\citep{li2022qrs_companion,li2023iet_faultprevention}.
LLM synthesis accelerates coding~\citep{chen2021codex,openai2023gpt4} but
does not respect guard order~\citep{dakhel2023copilot}.

\paragraph{Verifier-in-the-loop.}
DafnyBench~\citep{loughman2024dafnybench} evaluates LLM synthesis under full
deductive verification, repairing from compiler messages without FSF ordering
or PDR/FAR.
Clover~\citep{sun2024clover} is the closest CEGIS-style verifier loop for
Dafny; it lacks first-match SpecIR regions, typed scenario feedback, and a
pattern gate.
Our B6 (VerifierLoop-FSF) adapts witness-in-the-loop repair to FSF without
Semantic IR or hard pattern screening---a controlled comparator, not a
drop-in Clover clone.

\paragraph{Verbal / test repair loops.}
Self-Refine~\citep{madaan2023selfrefine}, Self-Debug~\citep{chen2023selfdebug},
and Reflexion~\citep{shinn2023reflexion} iterate on self-critique, execution
traces, or verbal memory.
They improve generation success but do not emit region-typed failure records
or measure prevention under conjunctive release.
Our B3--B5 instantiate those styles on the same BENCH-120 budget ($K{=}3$);
full numbers sit in the supplementary material.

\paragraph{Positioning.}
HSP-Agile sits below full deductive verification: bounded SMT
witnesses~\citep{demoura2008z3,barrett2018smt} and pattern guards approximate
release quality at agile latency~\citep{woodcock2009formalsurvey}.
Relative to prior work we contribute (i)~first-match SpecIR witnesses,
(ii)~typed Semantic Feedback IR as the repair hinge, and (iii)~conjunctive
$\mathrm{Accept}$ with PDR/FAR---evaluated against B1/B2 and ablations A1/A3
in the main text (B3--B6 in supp.).
E18 shows that naive witness{+}IR hybrids (\texttt{M\_lite}) do not recover
E6-level gains without the integrated gate.
