\subsection{Mechanism: Semantic Feedback IR (E6)}
Table~\ref{tab:e6-conf} holds $K$ and task count fixed and sweeps feedback
content under full~M ($K{=}3$).
Full Semantic IR (variant~C) improves mean conformance by
\textbf{+7.7\,pp} over unstructured test-only (A)---the paper's lead mechanism
claim.
Paired per-task: wins/losses/ties \textbf{14/4/102}; bootstrap 95\% CI on
$\Delta$Conf.\ $[2.3,\,13.6]$\,pp; Wilcoxon $p{=}0.018$
(Table~\ref{tab:e6-paired-conf}).
E14 (length-matched): \texttt{semantic\_ir} 75.1\% vs.\
\texttt{execution\_trace\_matched} 85.4\%---so +7.7\,pp is scoped to
unstructured test-only under~M, not uniqueness vs.\ any structured trace.

\begin{table}[t]
  \centering
  \caption{E6 feedback variants ($n{=}120$, $K{=}3$).}
  \label{tab:e6-conf}
  \footnotesize
  \begin{tabular}{lcc}
    \toprule
    Variant & Feedback & Conf.\ (\%) \\
    \midrule
    A & test only        & 79.1 \\
    B & test + expected  & 78.9 \\
    C & full Semantic IR & \textbf{86.9} \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[t]
  \centering
  \caption{E6 paired statistics (C vs.\ A/B).}
  \label{tab:e6-paired-conf}
  \footnotesize
  \begin{tabular}{lccc}
    \toprule
    Contrast & W/L/T & $\Delta$ (pp) & 95\% CI \\
    \midrule
    C$-$A & 14/4/102 & $+7.7$ & $[2.3,\,13.6]$ \\
    C$-$B & 14/4/102 & $+8.0$ & $[2.2,\,14.0]$ \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Prevention (E2)}
M improves prevention: PDR 95.0\%, FAR 5.0\% vs.\ B2 91.2\%\,/\,8.8\%
(impl-screening $n{=}852$; paired bootstrap 95\% CI on $\Delta$PDR excludes~0).
Escalate from B2 to~M when release quality is measured by escaped defects,
not mean Conf.\ alone.
Pattern-guard value is measured here (and E17 advisory), not by mean Conf.\
ablation under an advisory gate.

\subsection{Deployment boundary (C4)}
Canonical E3 tertiles are \emph{non-monotone} (M does not uniformly beat B2
by overlap alone)---\textbf{do not} escalate on tertiles.
\textbf{Decision rule:} prefer B2 by default; escalate to~M iff release
requires $\mathrm{Accept}{=}1$ \emph{or} target FAR${\le}5\%$ (E2 operating
point).
Industrial-pattern SOFL ($n{=}31$): B2 reaches 100\% Conf.\ under gpt-4o and
Claude (cross-model subsection).

\begin{table}[t]
  \centering
  \caption{Deployment boundary: when to prefer B2 vs.\ full M.}
  \label{tab:deploy}
  \footnotesize
  \begin{tabular}{@{}p{0.22\linewidth}p{0.34\linewidth}p{0.34\linewidth}@{}}
    \toprule
    \textbf{Signal} & \textbf{Prefer B2} & \textbf{Prefer M} \\
    \midrule
    Default / industrial & Mean Conf.\ / latency & --- \\
    Typed repair (E6) & --- & +7.7\,pp vs.\ test-only \\
    Release bar & Mean Conf.\ sufficient & Conjunctive $\mathrm{Accept}$ \\
    Prevention & --- & FAR${\le}5\%$ / PDR 95\% (E2) \\
    Overlap tertiles & Non-monotone (E3) & Do not use alone \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Ceiling stress-tests and equal-$K$ (E1)}
Equal-$K$ Conf.\ ranking (\texttt{run\_e1\_equal\_k\_v1}, both $K{=}3$):
\texttt{M\_eq} 100.0\% vs.\ B2 97.5\% (${+}2.5$\,pp; wins 3/120, losses~0,
ties~117).
Strengthened E1 bundle ($K{=}5$): M 100.0\% vs.\ B2 98.3\% (${+}1.7$\,pp;
wins 2/120)---stress-test only, not single-factor C2.
E10: ${+}1.0$\,pp; E12: B2 100.0\% / M 99.7\% (B2 first every seed).
\textbf{Do not claim M dominates all seeds or corpora.}
Lead Conf.\ mechanism remains E6 (+7.7\,pp with CI excluding~0).
