\chapter{Introduction}
\label{ch:intro}

A release dashboard can look healthy while a residual specification region is
already broken.
That is the defect this report studies---not whether a model can write more
tests that pass, but whether a team can \emph{refuse} to ship LLM-generated
code when the specification still has a thin slice that everyday suites never
exercise.

\textbf{Acceptance safety} is that refusal rule.
Release requires every specification-derived witness to match \emph{and} a
structural screen to clear:
$\mathrm{Accept}\Leftrightarrow\mathrm{Conf}{=}1\wedge\mathrm{Screen}{=}\mathit{pass}$.
We refer to this directly as a \textbf{dual-gate acceptance check}.
A candidate that ``looks right'' on busy inputs, yet mishandles a catch-all or
overlap region, must not sail through on mean pass rate alone.
HSP-Agile realises the check on SOFL/FSF through SpecIR regions,
structured scenario-aware repair feedback, and a configurable
$\mathrm{Screen}$.
The check is a bounded operational criterion evaluated over generated witnesses
and the configured screen; it is not a proof of whole-program safety or a new
SMT decision procedure.

The contribution is headroom-conditioned catch-all repair under a SpecIR
first-match oracle, plus a conditional dual-gate acceptance check.
Conjunctive Accept cuts escaped defects at comparable mean conformance (E2);
structured feedback rescues catch-all failures when repair still has headroom
(E6); prefer test-feedback (B2) by default and escalate to full mode~M when
release requires $\mathrm{Accept}{=}1$ or a contractual FAR budget exists
\emph{and} repair headroom remains.
Screen is a conditional second gate: it cuts FAR on industrial-style mutants
(56.8\% to 32.6\%) but does not attain FAR${\le}5\%$ on that slice.
The controlled causal estimate for SIFR is E6 on BENCH-120; Gemini hard-seed
and public-corpus runs bound saturation and transfer, not a second
${+}7.7$\,pp headline.
When test-only Conf.\ is already partial ($0{<}\mathrm{Conf}{<}1$), keep B2
rather than forcing Semantic IR.
When B2 already achieves Accept${=}1$ (or 100\% mean Conf.\ on the release
suite), do not escalate to~M---the fuller loop can regress under strong models.


\section{Primary Vignette: Catch-All That Ships}
\label{sec:intro:running-example}

Start from the failure mode that drives the strongest Conf.\ evidence.

Imagine a tier classifier for incoming support tickets
(Listing~\ref{lst:others-fsf}).
Two named guards handle the easy cases: reject negative scores; pass scores at
or above a floor.
Everything else falls into \texttt{OTHERWISE} and must return
\texttt{Review}---a human queue, not an automatic green light.
An LLM candidate that looks ``logically smooth'' collapses that residual into
\texttt{Pass} (Listing~\ref{lst:others-bug}).
Unit tests drawn from Reject and Pass bands all succeed.
The dashboard reports a high pass rate.
In production, the mid-band---the catch-all---never reaches a human; the bug
is an escaped defect, not a failed build.

That is false acceptance in miniature.
A pass/fail dump tells the repair model only that ``something failed'' if a
rare mid-band test ever appears; it does not name scenario~$S_3$
(\texttt{others}), the expected constant, or why the residual matters.
A typed feedback record does.
Under controlled E6, all 14 Conf.\ wins for the structured scenario-aware
feedback record (implemented as Semantic Feedback IR) over
test-only concentrate on this catch-all regime
(Tables~\ref{tab:e6-win-profile},~\ref{tab:e6-win-loss},~\ref{tab:e6-qual});
none are ordering-precedence repairs on BENCH-120.

\inputfsf{listings/others_rescue.fsf}{Primary vignette: catch-all
  \texttt{others} must return \texttt{Review}. Busy Reject/Pass tests miss
  the residual---the E6-aligned repair regime.}{lst:others-fsf}
\inputpython{listings/others_rescue_bug.py}{Looks coherent, ships the wrong
  catch-all: mid-band returns \texttt{Pass} instead of \texttt{Review}.}{lst:others-bug}


\section{Problem Statement}
\label{sec:intro:problem}

\subsection{Setting}
SOFL \citep{liu1998sofl,liu2004soflbook} provides the Functional Specification
Format (FSF): an ordered list of scenarios, each pairing a guard with a
postcondition under first-match precedence.
Agile-SOFL \citep{li2022qrs_companion} embeds those artefacts in sprint
workflows as design documents, oracles, and acceptance criteria.
LLM-assisted generation accelerates that workflow
\cite{chen2021codex,openai2023gpt4}, but synthesises by pattern completion
rather than by respecting specification structure
\cite{dakhel2023copilot,sun2024clover}.
Two failure modes matter: (i)~catch-all / output mismatches on residual
regions that unstructured dumps fail to localise (primary vignette above),
and (ii)~overlap slices where first-match order is decisive (secondary
vignette below).
Random suites undersample both \citep{zhu1997specification}; when they fail,
pass/fail bits still omit scenario identity.

\subsection{High Pass Rate, Weak Acceptance Safety}
A specification-conformance defect is a bug that unit tests rarely catch---or
catch without enough structure to repair.
On the tier classifier, a candidate can clear every Reject/Pass sample and
still mis-route every mid-band ticket.
A release gate that only looks at mean pass rate may \emph{accept} that
candidate anyway.

We write $\mathrm{FAR}$ for the \textbf{False Accept Rate}: among injected
faulty implementations (mutants), the fraction that a mode incorrectly
releases.
$\mathrm{PDR}$ (Prevention Detection Rate) is the complementary rejection
rate on that set.
Strong mean conformance---agreement on a finite witness suite---does not
imply low FAR.

\subsection{Acceptance Predicate}
Given task~$t$ with ordered scenarios $\mathcal{S}_t = (S_1,\ldots,S_n)$ and
budget~$K$, is candidate~$c$ safe to release?
Every specification-derived witness must match, \emph{and} no critical
structural anti-pattern may remain:
\begin{equation}
  \mathrm{Accept}(c, t)
  \;\iff\;
  \mathrm{Conf}(c, t) \geq \tau
  \;\wedge\;
  P^H(c) = 0,
  \label{eq:acceptance}
\end{equation}
where $\mathrm{Conf}(c,t)\in[0,1]$ is the fraction of witnesses on which $c$
matches the oracle, $\tau=1.0$ here, and $P^H(c)=0$ asserts that the
pluggable high-severity $\mathrm{Screen}$ is clear.
A witness for scenario~$S_i$ must land in the \emph{priority region}
\begin{equation}
  \Phi_i(\mathbf{x})
  \;\triangleq\;
  G_i(\mathbf{x}) \wedge \bigwedge_{j<i}\neg G_j(\mathbf{x})
  \label{eq:phi-intro}
\end{equation}
---including, for catch-alls, an input that satisfies no earlier guard.
Figure~\ref{fig:defect-vs-test} contrasts a busy test cloud with a surgical
SMT cut; the geometry applies to residual regions as well as overlaps.

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/defect_vs_test}
  \caption{\textbf{Acceptance safety needs witnesses, not denser random tests.}
    Left: unit tests concentrate on busy bands, so a catch-all or overlap bug
    can still look like a high pass rate.
    Right: an SMT model of a thin $\Phi_i$ (Equation~\eqref{eq:phi-intro})
    forces the region the release gate must clear---why FAR, not mean Conf.\
    alone, is the release metric.
    Catch-all rescue under structured feedback is the primary E6 Conf.\
    mechanism (Listings~\ref{lst:others-fsf}--\ref{lst:others-bug});
    ordering/overlap motivates witnesses and E2.}
  \label{fig:defect-vs-test}
\end{figure}

\subsection{Secondary Vignette: Ordering / Overlap}
Overlap explains a second way tests miss bugs and motivates $\Phi_i$ witnesses.
Listing~\ref{lst:tax-fsf} is an ordered tax-bracket contract: on
$(\mathit{income},\mathit{is\_foreign})=(150000,1)$, both a foreign-high
surcharge and a domestic mid bracket hold, yet first-match requires
\texttt{ForeignHigh}.
Listing~\ref{lst:tax-bug} tests mid first---wrong, but often invisible to
domestic-heavy suites.
This vignette anchors witness design and E2 structural mutants; it is
\emph{not} the dominant E6 win regime
(Table~\ref{tab:defect-evidence-map}).

\inputfsf{listings/compute_tax.fsf}{Secondary vignette: ordered tax brackets.
  First-match: $S_2$ preempts $S_3$ on the foreign-high overlap.}{lst:tax-fsf}
\inputpython{listings/compute_tax_bug.py}{Bracket-order bug: mid guard before
  foreign-high surcharge---witness/E2 motivation, not E6 win coding.}{lst:tax-bug}


\section{Research Challenges}
\label{sec:intro:challenges}

\subsection{Name the Failed Region}
Checking scenario~$S_i$ requires an input in $\Phi_i$
(Equation~\eqref{eq:phi-intro})---including catch-alls that random tests miss
\cite{barrett2018smt}.
The check must preserve first-match partitions without rewriting the pipeline
per surface notation (Chapter~\ref{ch:formalization}).

\subsection{Typed Failures for Repair}
Test-feedback loops lack specification-level structure
\cite{sun2024clover}: the model may patch one counterexample without knowing
whether a catch-all output or an ordering clash failed.
Repair needs a record that names the violated scenario \emph{before} any
natural-language prompt is rendered (Chapter~\ref{ch:method}).

\subsection{Conjunctive Release under Budget~$K$}
Full verification is impractical for unannotated LLM output
\cite{leino2010dafny,loughridge2024dafnybench}.
What is needed is calibrated acceptance safety: stronger than pass-rate gates
\cite{chen2021codex}, weaker than deductive proof, and fast enough for sprint
budgets \citep{beck2004xp,li2022qrs_companion}---Equation~\eqref{eq:acceptance}
under hard~$K$ ($K{=}3$ in controlled E6).


\section{Proposed Approach}
\label{sec:intro:approach}

The same specification that exposes the catch-all defect can drive prevention
if three ideas hold: shared semantic regions, typed repair feedback, and
conjunctive release.
\textsc{SgDP} (Specification-guided Defect Prevention) does three jobs:
(1)~carve first-match regions with an SMT solver; (2)~tell the LLM
\emph{which} region failed; (3)~refuse release unless every region and the
pluggable screen clear.
Figure~\ref{fig:sgdp-overview} is that loop.
Adapters and SMT lowering are necessary engineering, not separate scientific
claims.
Chapter~\ref{ch:method} walks the decision path end to end;
Chapter~\ref{ch:results} leads with mechanism and prevention, not aggregate
pass-rate crowns.

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/sgdp_framework_overview}
  \caption{\textbf{Three ideas, one release loop:} SpecIR regions, Semantic
    Feedback IR, conjunctive $\mathrm{Accept}$ under budget~$K$.
    On the catch-all vignette, the loop forces the residual $\Phi_i$, names
    the others/output violation, and releases only when Accept holds
    (Equation~\eqref{eq:acceptance}).
    Left to right for one attempt; dashed orange arc is repair; dashed branch
    is FSF lowering and the pluggable pattern screen.}
  \label{fig:sgdp-overview}
\end{figure}


\section{Contributions}
\label{sec:intro:contributions}

This work contributes an \textbf{operational acceptance-checking stack} for
ordered guards---a SpecIR witness interface, headroom-conditioned catch-all
repair, and a conditional dual-gate release metric---not a new SMT solver or a
proof system (Figure~\ref{fig:contribution-map};
Table~\ref{tab:evidence-ladder}).
The headline is acceptance safety (C3): Screen is a conditional second gate
that separates witness-conformant candidates only when industrial-pattern
mutants exercise it.
SpecIR adapters, SMT lowering, and the hard benchmark are engineering that
makes the decision rule runnable.
As a formal-engineering method, the dual-gate check is a bounded operational
criterion under a faithful SpecIR encoding, not a theorem of program safety.

\input{tables/evidence_ladder}

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/contribution_map}
  \caption{\textbf{Tier-1 claims are C2--C4}, with acceptance safety (C3) as
    the release headline.
    SpecIR is the shared engineering hinge, not a monotone-overlap ranking
    claim.
    C3 gates release; C2 enables rescue under headroom; C4 chooses when to
    pay for the full loop.}
  \label{fig:contribution-map}
\end{figure}

\paragraph{Engineering premise (SpecIR).}
Prevention must depend on \emph{semantic regions}, not surface syntax---the
$\Phi_i$ cut in Figure~\ref{fig:defect-vs-test}.
Overlap tertiles on E3 are non-monotone: denser overlap does not by itself
prescribe deploying M (a negative bound on C4).

\begin{enumerate}
  \item[\textbf{C3.}]
    Release must be \emph{conjunctive}: every specification-derived witness
    matches \emph{and} no high-severity structural pattern remains
    (Equation~\eqref{eq:acceptance}).
    Pass-rate alone is not enough---as the catch-all vignette shows, a
    candidate can look strong on busy tests and still ship a broken residual.
    On industrial-style mutants ($n{=}132$), FAR falls from 56.8\% to
    32.6\% under M---the primary prevention claim (bootstrap 95\% CI on the
    FAR drop: $[-31.8,\,-17.4]$\,pp; excludes~0).
    Pooled over $n{=}852$ that lift appears as M PDR 95.0\% / FAR 5.0\%
    vs.\ B2 91.2\% / 8.8\%, but HardSynthetic mutants are already saturated
    and do not exercise Screen; pooled FAR 5.0\% is \emph{not} the release
    claim.
    The HardSynthetic slice ($n{=}720$) already saturates at PDR 100\% /
    FAR 0\%.
    Ordering and preemption mutants motivate the gate beyond E6 Conf.\ gains.
    Limitation: near-ceiling mean-Conf.\ ranking across corpora is secondary
    appendix material; C3 is scored by Accept/FAR, not by Conf.\ crowns.

  \item[\textbf{C2.}]
    When repair has headroom, it needs \emph{specification-indexed structured
    feedback}, not raw pass/fail bits.
    E6 isolates feedback under full~M, equal $K{=}3$: semantic\_ir reaches
    86.9\% vs.\ test\_only 79.1\% (${+}7.7$\,pp; paired W/L/T $14/4/102$;
    95\% CI $[2.3,\,13.6]$\,pp; Wilcoxon $p{=}0.018$).
    The primary reading is headroom-conditional: on tasks where test-only
    Conf.\ is already~$0$ ($n{=}13$), semantic\_ir rescues all cases
    (${+}90.7$\,pp; 13/0/0); on the broader headroom set (test-only
    Conf.${<}1$, $n{=}114$), the paired mean is ${+}8.2$\,pp (95\% CI
    excludes~0).
    The all-task ${+}7.7$\,pp aggregate is a supporting summary, not a
    universal repair claim.
    Among the 18 decisive (non-tie) tasks, Semantic IR wins 14/18 (77.8\%;
    Wilson 95\% CI $[54.8,\,91.0]$).
    The partial-failure band ($0{<}$test-only Conf.${<}1$) does not favour
    typed IR (${-}2.5$\,pp); keep B2 rather than forcing Semantic IR there.
    Qualitative coding of all 14 wins (Table~\ref{tab:e6-qual}):
    \textbf{14/14 catch-all (\texttt{others}) rescues};
    \textbf{0/14 ordering-precedence repairs}.
    Field structure (Table~\ref{tab:field-structure}): FULL beats NL-only
    dumps of the same facts (${+}28.0$\,pp; CI excludes~0).
    Limitation: E14 shows length-matched structured traces can beat bare
    Semantic Feedback IR on aggregate Conf.; the E6 win profile on BENCH-120
    is catch-all rescue, not ordering-precedence repair.

  \item[\textbf{C4.}]
    Prefer test-feedback (B2) by default for mean conformance and latency;
    escalate to full HSP-Agile (M) when release requires
    $\mathrm{Accept}{=}1$ or a contractual FAR budget exists \emph{and}
    repair headroom remains (Table~\ref{tab:deployment-boundary}).
    Screen helps on industrial-pattern mutants but does not attain
    FAR${\le}5\%$ on that slice.
    When contracts are simple enough that baselines already saturate, B2 is
    enough---that is a deployment boundary, not a failure of the decision
    rule.
    Equal-$K$ \texttt{M\_eq} is ${+}2.5$\,pp vs.\ B2 ($100.0\%$ vs.\ $97.5\%$;
    $3/0/117$); external pilots often tie on mean Conf.; a public ACL-style
    case (Section~\ref{sec:results:acl-case}) shows the gate intercepting a
    lucky-pass candidate.
    A pre-hoc escalation checklist (HTE) is a deployment policy, not a
    validated predictor or a fourth algorithmic contribution (precision
    0.167 on hard-task metadata).
    Limitation: live vendor CI replication remains future work; present
    pilots support the deployment heuristic, not production KPIs.
\end{enumerate}

Primary evidence is E2, E6, equal-$K$ Conf.\ ranking, and the deployment table.
We evaluate mechanism (E6), prevention (E2), and saturation boundaries under
stronger models or public corpora as separate evidence tracks.
Near-ceiling fixed-oracle ranking runs (E1/E10/E12) are appendix material:
E1 places B1/B2/M at $84.2\%/98.3\%/100.0\%$ under a $K{=}5$ bundle for M;
E12 shows B2 at $100\%$ every seed and M at $99.7\%$---M does not lead every
seed.


\section{Research Questions}
\label{sec:intro:rq}

Evaluation is mechanism- and safety-first; aggregate ranking is secondary.

\textbf{RQ1}---does specification-indexed Semantic Feedback IR improve repair
beyond unstructured test-only feedback? (E6; qualitative win coding; field
probe);
\textbf{RQ2}---does the M--baseline gap grow monotonically with guard-overlap
density? (E3, E10; negative bound on C4);
\textbf{RQ3}---does conjunctive $\mathrm{Accept}$ improve acceptance safety
(PDR/FAR) vs.\ mean conformance alone? (E2);
\textbf{RQ4}---under what conditions should practitioners deploy M vs.\ B2?
(deployment boundary; equal-$K$; ACL case);
\textbf{RQ5}---quality--overhead trade-off (appendix ranking runs at
comparable Conf.).


\section{Report Organisation}
\label{sec:intro:organisation}

Chapter~\ref{ch:background} surveys background and related work (including
LLM-as-judge structured feedback).
Chapter~\ref{ch:formalization} develops the mathematical core.
Chapter~\ref{ch:method} designs SpecIR, typed IR, and the pluggable Screen.
Chapter~\ref{ch:implementation} is supporting engineering (Screen
pluggability over module taxonomy).
Chapters~\ref{ch:experiments}--\ref{ch:results} lead with E6, E2, equal-$K$,
and C4; near-ceiling corpora appear in the appendix.
Chapter~\ref{ch:discussion} states scope when baselines saturate;
Chapter~\ref{ch:conclusion} restates the decision framework.
Appendices cover reproducibility, benchmark construction, the pattern
catalogue, and near-ceiling ranking runs (Appendix~\ref{app:stress}).
