\chapter{Formal Problem Definition}
\label{ch:formalization}

Chapter~\ref{ch:intro} stated the prevention problem in words.
Making that problem precise is the job of this chapter.
The sections below fix what a task is, how ordered guards are evaluated, how
conformance and acceptance are scored, and how the hard benchmark is
characterised---always returning to \code{classify\_signal} so the notation
never floats free of an example.

Notation used throughout: $\mathbb{N}$ for the non-negative integers,
$\mathbb{Z}$ for all integers, and $\mathbf{1}[\cdot]$ for the Iverson
bracket ($1$ when its argument is true and $0$ otherwise).

% ---------------------------------------------------------------------------
\section{Running Example Roadmap}
\label{sec:form:roadmap}
% ---------------------------------------------------------------------------

Readers should keep one concrete task in mind while reading the definitions.
That task is \code{classify\_signal(level, threshold)} from
Chapter~\ref{ch:intro}: return \texttt{"Error"} when the level is negative,
\texttt{"Critical"} when the level meets or exceeds a threshold (and is
non-negative), and \texttt{"Normal"} otherwise.
Each formal object is followed by a short \emph{Example} paragraph that
instantiates it on this task.
Figure~\ref{fig:specir-walkthrough} previews the path from FSF text through an
intermediate representation and a priority formula to a concrete witness.

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/specir_walkthrough}
  \caption{\textbf{From FSF text to witness (classify\_signal).}
    The adapter lowers surface syntax into SpecIR; the witness generator
    encodes first-match priority as $\Phi_i$; Z3 returns a model such as
    $(-3,-10)$ that activates $S_1$.
    Takeaway: SpecIR is the hinge between notation-specific text and the
    shared prevention pipeline.}
  \label{fig:specir-walkthrough}
\end{figure}

% ---------------------------------------------------------------------------
\section{Formal Task Model}
\label{sec:form:task}
% ---------------------------------------------------------------------------

Release decisions need a precise unit of work: what must the pipeline know
about a sprint artefact before it can judge a candidate?
That unit is a \emph{formal specification task}---a single function whose
behaviour is described entirely by an ordered scenario specification.
HSP-Agile instantiates the model over SOFL/FSF~\citep{li2022agilesofl,liu2014sofl};
every later stage consumes one such task.

\begin{definition}[Formal Specification Task]
\label{def:task}
A \emph{formal specification task} is a triple
\[
  t \;=\; \bigl(\mathcal{S}_t,\;\mathcal{X}_t,\;\mathcal{Y}_t\bigr),
\]
where
\begin{itemize}
  \item $\mathcal{S}_t = \langle s_1, s_2, \ldots, s_n \rangle$ is a finite,
        ordered sequence of $n$ FSF scenarios (the \emph{specification});
  \item $\mathcal{X}_t \subseteq \mathbb{Z}^d$ is the \emph{input domain},
        a $d$-dimensional integer vector space; and
  \item $\mathcal{Y}_t \subseteq \mathbb{Z}^m$ is the \emph{output domain},
        an $m$-dimensional integer vector space.
\end{itemize}
\end{definition}

\paragraph{Example.}
For \code{classify\_signal}, $d=2$, $m=1$ (string-valued alert encoded as a
discrete label), $\mathcal{S}_t=\langle s_1,s_2,s_3\rangle$, and
$\mathcal{X}_t\subseteq\mathbb{Z}^2$ with coordinates
$(\mathit{level},\mathit{threshold})$.
Benchmark tasks in later chapters often use $d=5$ and $m=3$; the same
definitions apply.

% ---------------------------------------------------------------------------
\section{Specification Intermediate Representation (SpecIR)}
\label{sec:form:specir}
% ---------------------------------------------------------------------------

The prevention pipeline should not be rewritten for every surface notation.
What is needed is a shared hinge: a notation-agnostic form that still preserves
ordered guarded cases.
That hinge is the \textbf{Specification Intermediate Representation}
(\textbf{SpecIR}).
Adapters map SOFL/FSF, Mini-Z, or Mini-StateMachine text into SpecIR; the rest
of the pipeline depends only on SpecIR.

\begin{definition}[Specification Intermediate Representation]
\label{def:specir}
For task $t$, $\text{SpecIR}(t) = (\Sigma_t,\; \Gamma_t,\; C_t,\; M_t)$, where
$\Sigma_t$ is the typed I/O signature, $\Gamma_t$ the variable domain,
$C_t = \langle c_1,\ldots,c_n\rangle$ an ordered list of guarded cases
(first-match semantics), and $M_t$ notation-specific metadata.
The abstract syntax is written
$\text{SpecIR} \triangleq \langle\Sigma_t, \Gamma_t, C_t, M_t\rangle$ or,
equivalently, $\text{SpecIR} ::= \texttt{Signature}\;\texttt{Cases}\;[\texttt{Metadata}]$.
\end{definition}

\paragraph{Example.}
For \code{classify\_signal}, $\Sigma_t$ declares inputs
\texttt{level:int}, \texttt{threshold:int} and output \texttt{alert:string};
$C_t$ holds three guarded cases whose first guard atom is
$\{\texttt{var}{=}\texttt{level},\;\texttt{op}{=}\texttt{lt},\;
\texttt{threshold}{=}0\}$ (Figure~\ref{fig:specir-walkthrough}).

Adapters and validators need a concrete surface form.
The EBNF below is that form: it matches the JSON schema used in the
implementation, so every SpecIR document the pipeline accepts can be checked
mechanically.

\begin{definition}[SpecIR Surface Grammar (EBNF)]
\label{def:specir-ebnf}
The JSON serialisation aligns with \texttt{schemas/spec\_ir.schema.json} and
\texttt{src/ir/spec\_ir.py}:
\begin{lstlisting}[language={},basicstyle=\ttfamily\footnotesize]
SpecIR      ::= task_id notation name Signature Cases surface_prompt [Metadata]
Signature   ::= Inputs Outputs
Inputs      ::= Param*          Outputs ::= Param*
Param       ::= name type       (* nat | int | bool | string *)
Cases       ::= GuardedCase+
GuardedCase ::= index guard postcondition [guard_text] [post_text]
Guard       ::= GuardAtom* | { var="" op="others" }
GuardAtom   ::= var op [threshold]   (* op in eq,ne,lt,le,gt,ge,others *)
\end{lstlisting}
\end{definition}

Not every string that parses is a usable specification.
Well-formedness rules out missing variables, non-consecutive indices, and a
final case that fails to cover the residual input space.

\begin{definition}[SpecIR Well-formedness]
\label{def:specir-wf}
A SpecIR document $d$ is \emph{well-formed} iff:
(i)~$d.\texttt{notation}$ is one of \texttt{sofl}, \texttt{mini\_z},
\texttt{mini\_statemachine}, or \texttt{real\_derived};
(ii)~case indices are consecutive $1,\ldots,n$ with pairwise distinct guards;
(iii)~every \texttt{guard} atom's \texttt{var} is declared in \texttt{inputs},
    except the \texttt{others} sentinel;
(iv)~every \texttt{postcondition} key is declared in \texttt{outputs}; and
(v)~the final case is either an \texttt{others} sentinel or its guard is
    logically equivalent to $\neg\bigvee_{i<n} g_i$.
Validation is enforced by \texttt{src/ir/schema\_validate.py}
(Appendix~\ref{sec:repro:schemas}).
\end{definition}

Surface notations still differ.
An adapter is the thin translation that turns notation-specific text into
well-formed SpecIR; after that step, the shared pipeline never sees the
original syntax.

\begin{definition}[Specification Adapter]
\label{def:spec-adapter}
A \emph{Specification Adapter} is $\text{Adapt}_L : \text{Text}_L \longrightarrow \text{SpecIR}$.
\end{definition}

HSP-Agile optionally projects SpecIR back onto an FSF-shaped view for
screening that still speaks in scenarios.
That projection is lowering; the core stages continue to depend only on SpecIR.

\begin{definition}[FSF Lowering (HSP-Agile Instantiation)]
\label{def:fsf-lowering}
$\text{Lower}_{\text{FSF}} : \text{SpecIR} \longrightarrow \text{TaskSpec}_{\text{FSF}}$
projects SpecIR onto the FSF view; the pipeline core depends only on SpecIR.
\end{definition}

Each ordered case in SpecIR corresponds to one FSF scenario: a guard that
selects a region of the input space, and an output mapping that says what
must be returned there.

\begin{definition}[FSF Scenario]
\label{def:scenario}
A \emph{Functional Scenario Framework (FSF) scenario} is a pair
\[
  s_i \;=\; (g_i,\;\phi_i),
\]
where
\begin{itemize}
  \item $g_i : \mathcal{X}_t \to \{0, 1\}$ is a Boolean \emph{guard
        condition} over the input domain; and
  \item $\phi_i : \mathcal{X}_t \to \mathcal{Y}_t$ is a total
        \emph{output mapping} that prescribes the correct output for every
        input satisfying the guard.
\end{itemize}
\end{definition}

\paragraph{Example.}
$s_1=(\mathit{level}<0,\;\phi_1)$ with $\phi_1\equiv\texttt{"Error"}$;
$s_2=(\mathit{level}\ge\mathit{threshold},\;\phi_2)$ with
$\phi_2\equiv\texttt{"Critical"}$; $s_3$ is the \texttt{others} case with
$\phi_3\equiv\texttt{"Normal"}$.

Guards are expressed as linear arithmetic predicates over the integer
variables of $\mathcal{X}_t$.  In practice they encode conditions such as
threshold comparisons ($a > \theta_1$), conjunctions of comparisons, or
negations thereof.  Output mappings specify concrete assignments to the output
variables, making each scenario a self-contained input-output contract.

\begin{definition}[Scenario Ordering and Specification Oracle]
\label{def:oracle}
Scenario $s_i$ has \emph{higher priority} than $s_j$ whenever $i < j$.
For any input $x \in \mathcal{X}_t$, the \emph{specification oracle}
$g_t : \mathcal{X}_t \to \mathcal{Y}_t$ is defined by first-match
evaluation:
\[
  g_t(x) \;=\; \phi_i(x),
  \qquad \text{where } i = \min\bigl\{j : g_j(x) = 1\bigr\}.
\]
The final scenario $s_n$ is the \emph{``others'' case} with guard
\[
  g_n(x) \;=\; \neg\,\bigvee_{j=1}^{n-1} g_j(x),
\]
ensuring that every input is handled by exactly one scenario.
\end{definition}

The first-match semantics above is the key property distinguishing FSF from
a simple case statement \citep{dijkstra1975guarded}: guard evaluation order
is part of the specification, not an implementation choice.
A candidate that evaluates $g_2$ before $g_1$ may be correct on most inputs
(where guards are disjoint) but wrong wherever both guards hold.

\paragraph{Example.}
On $(-3,-10)$, both $g_1$ and $g_2$ hold, so
$g_t(-3,-10)=\phi_1(-3,-10)=\texttt{"Error"}$.
An implementation that checks $g_2$ first returns \texttt{"Critical"} and
fails the oracle.

% ---------------------------------------------------------------------------
\section{FSF Scenario Semantics}
\label{sec:form:semantics}
% ---------------------------------------------------------------------------

The oracle in Definition~\ref{def:oracle} is a mathematical object.
The pipeline must make it operational: it needs concrete inputs on which to
execute candidates and compare outputs.
Those inputs are the concrete cases below, generated by SMT solving rather than
by random sampling.

\begin{definition}[Concrete Case]
\label{def:concrete-case}
A \emph{concrete case} for task $t$ is a pair
\[
  (x,\;y) \;=\; \bigl(x,\;g_t(x)\bigr),
  \qquad x \in \mathcal{X}_t.
\]
The set of all concrete cases is called the \emph{case set} $\mathcal{D}(t)$.
\end{definition}

In practice $\mathcal{X}_t$ is too large for exhaustive enumeration, so
$\mathcal{D}(t)$ is constructed by targeted SMT sampling \citep{demoura2008z3, barrett2018smt}:
one representative input per scenario.

\begin{definition}[SMT-Generated Test Suite]
\label{def:test-suite}
The case set $\mathcal{D}(t)$ is generated by the Z3 SMT solver as
follows.  For each scenario $s_i = (g_i, \phi_i)$, solve the satisfiability
query
\[
  x_i \;\models\; g_i(x) \;\land\; \bigwedge_{j < i} \neg\,g_j(x).
\]
If satisfiable, add $\bigl(x_i,\;g_t(x_i)\bigr)$ to $\mathcal{D}(t)$.
\end{definition}

\paragraph{Example.}
For $s_1$ the query is simply $\mathit{level}<0$; a model is $(-3,-10)$ with
oracle output \texttt{"Error"}.
For $s_2$ the query is
$\mathit{level}\ge\mathit{threshold}\land\neg(\mathit{level}<0)$; a model
might be $(5,3)$ with oracle output \texttt{"Critical"}.

The conjunct $\bigwedge_{j < i} \neg\,g_j(x)$ enforces strict scenario
membership: $x_i$ activates $s_i$ \emph{and no higher-priority scenario}.
This guarantees that the oracle maps $x_i$ through $\phi_i$ and not through
any $\phi_j$ with $j < i$, making each case a direct witness to the
prescribed output of exactly one scenario.

The following proposition establishes that the procedure achieves full
scenario coverage.

\begin{lemma}[Scenario Coverage]
\label{lem:coverage}
If the SMT solver finds satisfying inputs for all $n$ scenarios, then
$\mathcal{D}(t)$ contains at least one representative case per scenario,
covering the entire precedence ordering $s_1 \succ s_2 \succ \cdots \succ s_n$.
\end{lemma}

\begin{proof}
By construction, each $x_i$ satisfies $g_i$ but not $g_1, \ldots, g_{i-1}$.
Hence $i = \min\{j : g_j(x_i) = 1\}$, so the oracle evaluates
$g_t(x_i) = \phi_i(x_i)$.  Each distinct scenario $s_i$ therefore
contributes at least one case whose correct output is determined by
$\phi_i$, and the precedence ordering is witnessed by all $n$ cases jointly.
\end{proof}

Precedence defects concentrate near guard boundaries: inputs that sit close
to where one scenario yields to another.
That thin set is the boundary region; random tests rarely land inside it.

\begin{definition}[Boundary Region]
\label{def:boundary-region}
For margin $\varepsilon > 0$, the \emph{boundary region} is
$\mathcal{B}_\varepsilon(t) = \{\mathbf{x} \in \mathcal{X}_t :
  \exists i,\; \partial g_i(\mathbf{x}) < \varepsilon\}$.
\end{definition}

How often a faulty candidate fails on a randomly drawn witness is its
fault-exposure probability.
When that probability is small, unit-test repair loops look healthy while
ordering bugs remain.

\begin{definition}[Per-witness Fault-Exposure Probability]
\label{def:fault-exposure}
For faulty candidate $c$, $\varepsilon_c =
\Pr_{\mathbf{x} \sim \mathcal{D}(t)}[\,f_c(\mathbf{x}) \neq g_t(\mathbf{x})\,]$;
$\varepsilon = \min_{c \in \mathcal{C}\setminus\{g_t\}} \varepsilon_c > 0$.
\end{definition}

In plain terms: random tests miss the thin overlap slivers where ordering
bugs live; SMT witnesses are built to land there.
The corollary makes that contrast quantitative.

\begin{corollary}[Random vs.\ SMT Fault-Exposure Hit Rate]
\label{cor:random-vs-smt}
Let $\varepsilon = \mu(\mathcal{B}_\varepsilon(t))/\mu(\mathcal{X}_t)$ denote the
measure of the boundary region relative to the input domain.
For a faulty candidate $c$ with per-witness fault-exposure probability
$\varepsilon_c \geq \varepsilon$ (Definition~\ref{def:fault-exposure}),
a single uniform random sample detects the fault with probability at most
$\varepsilon_c$, because
$\Pr[\mathbf{x} \in \mathcal{B}_\varepsilon(t)] =
\mu(\mathcal{B}_\varepsilon(t))/\mu(\mathcal{X}_t)$ under uniform sampling---
typically small when guard overlap is dense.
The Constraint-guided Witness Generator produces one witness per strict
scenario region via $\Phi_i$ (Equation~\eqref{eq:priority}); under sound
encoding, each such witness activates scenario~$i$ with probability~$1$
(Theorem~\ref{thm:witness-soundness}), lying on (or within numeric precision of)
the boundary of~$g_i$.
Random sampling therefore reaches precedence-critical regions with probability
$\mathcal{O}(\varepsilon)$, whereas SMT achieves targeted boundary coverage
deterministically.
\end{corollary}

\begin{proof}
Assume faults of interest manifest only on inputs in
$\mathcal{B}_\varepsilon(t)$.
Under uniform sampling,
$\Pr[f_c(\mathbf{x})\neq g_t(\mathbf{x})]
  \le \Pr[\mathbf{x}\in\mathcal{B}_\varepsilon(t)]=\varepsilon$.
For SMT, satisfiability of $\Phi_i$ yields
$g_i(\mathbf{x}_i^*)\land\bigwedge_{j<i}\neg g_j(\mathbf{x}_i^*)$, so
$\mathbf{x}_i^*$ lies in the exclusive first-match region of~$s_i$ and
activates that region with probability~$1$ under a sound encoding
(Theorem~\ref{thm:witness-soundness}).
Hence random sampling hits precedence-critical regions with probability
$\mathcal{O}(\varepsilon)$, while SMT hits them deterministically.
\end{proof}

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/fsf_semantics}
  \caption{\textbf{FSF evaluation semantics for \code{classify\_signal}.}
    Left: the three-scenario FSF definition.
    Right: input-space partitioning with priority $S_1 \succ S_2 \succ S_3$;
    the green marked point $(-3,-10)$ is an $S_1$ witness that also lies in
    the geometric region of $S_2$'s guard.
    Takeaway: first-match, not geometric inclusion alone, decides the oracle.}
  \label{fig:fsf-semantics}
\end{figure}

The point of Figure~\ref{fig:fsf-semantics} is the green witness $(-3,-10)$:
geometrically it also satisfies $S_2$'s guard, yet first-match assigns it to
$S_1$.
SMT case generation places one such witness in each strict scenario region.

Pass/fail alone is not enough for repair.
The pipeline also needs a counterexample that names the violated scenario,
so the model can localise the bug inside the ordered specification.

\begin{definition}[Strict Scenario Violation]
\label{def:violation}
A candidate $c$ \emph{violates scenario $s_i$} on the concrete case
$(x_i, y_i) \in \mathcal{D}(t)$ if
\[
  f_c(x_i) \;\neq\; y_i \;=\; g_t(x_i),
\]
where $f_c : \mathcal{X}_t \to \mathcal{Y}_t$ denotes the observable
behaviour of $c$.  The associated \emph{counterexample} is the triple
\[
  (x_i,\;g_t(x_i),\;f_c(x_i)),
\]
comprising the failing input, the expected output, and the actual output
produced by $c$.
\end{definition}

\paragraph{Example.}
For Listing~\ref{lst:llm-incorrect} on $(-3,-10)$, the counterexample is
$\bigl((-3,-10),\;\texttt{Error},\;\texttt{Critical}\bigr)$ and the violated
scenario is $s_1$.

Counterexamples are the primary feedback signal returned to the LLM.
They are structured: the pipeline identifies not only that the candidate
produced a wrong answer, but \emph{which scenario} was active, allowing the
language model to localise the bug within the FSF specification.

% ---------------------------------------------------------------------------
\section{The Bounded Acceptance Problem}
\label{sec:form:acceptance}
% ---------------------------------------------------------------------------

The oracle and the case set answer whether a candidate is right on the
witnesses that matter.
Release, however, is a stronger decision: the candidate must clear those
witnesses \emph{and} a structural screen, and the search for such a candidate
must terminate within a fixed LLM budget.
The definitions below make that decision precise.

\begin{definition}[Candidate Implementation]
\label{def:candidate}
A \emph{candidate implementation} for task $t$ is any computable function
$c : \mathcal{X}_t \to \mathcal{Y}_t$.
Write $f_c$ for its observable input-output behaviour---the function computed
when $c$ is executed.
\end{definition}

A natural quality score is the fraction of witnesses on which the candidate
matches the oracle.
That score is \emph{strict formal conformance}: it is continuous, so partial
progress during repair remains visible, and it is grounded in the
specification rather than in an arbitrary unit-test suite.

\begin{definition}[Strict Formal Conformance]
\label{def:conformance}
The \emph{strict formal conformance} of candidate $c$ with respect to
task $t$ is
\[
  \mathrm{Conf}(c,\,t)
  \;=\;
  \frac{1}{|\mathcal{D}(t)|}
  \sum_{(x,\,y)\,\in\,\mathcal{D}(t)}
  \mathbf{1}\bigl[f_c(x) = y\bigr].
\]
$\mathrm{Conf}(c,t) \in [0, 1]$; a value of $1$ indicates that $c$ agrees
with the oracle on every generated case.
\end{definition}

\paragraph{Example.}
If $\mathcal{D}(t)$ has three witnesses and the buggy
\code{classify\_signal} fails only on $(-3,-10)$, then
$\mathrm{Conf}(c,t)=2/3$.

\begin{definition}[Strict Success]
\label{def:strict-success}
Candidate $c$ \emph{strictly succeeds} on task $t$ if and only if
\[
  \mathrm{Conf}(c,\,t) \;\geq\; \tau,
  \qquad \tau = 1.
\]
\end{definition}

The choice $\tau = 1$ (all generated cases must pass) is deliberate.
A threshold below $1$ would tolerate partial conformance and mask
scenario-ordering violations: a candidate with wrong guard order may pass
most cases where overlapping guards prescribe identical outputs, yet fail
systematically where they disagree \citep{zhu1997specification}.
Hence all $n$ generated cases must pass before formal conformance is declared.

Conformance alone is insufficient for acceptance; the pipeline also screens
candidates for the presence of high-severity requirement fault patterns.

\begin{definition}[High-Severity Pattern Violation]
\label{def:pattern-violation}
Let $\mathcal{P}^H$ be the set of high-severity requirement fault patterns
(as enumerated in Appendix~\ref{app:patterns}).  Candidate $c$ is
\emph{pattern-safe} with respect to task $t$ if
\[
  \forall p \in \mathcal{P}^H : p(c,\,t) = 0.
\]
Each pattern checker $p : (c, t) \mapsto \{0, 1\}$ returns $1$ when the
pattern is detected.
\end{definition}

$\mathcal{P}^H$ contains the \textbf{nine} high-severity patterns (severity = high or
critical in Table~\ref{tab:patterns}); the remaining five patterns carry medium or low
severity and inform repair feedback without blocking acceptance.
Each pattern is encoded as a static analysis rule applied to the candidate source.

\begin{definition}[Hybrid Acceptance Predicate]
\label{def:accept}
The \emph{hybrid acceptance predicate} for candidate $c$ and task $t$ is
\[
  \mathrm{Accept}(c,\,t)
  \;=\;
  \mathbf{1}\!\left[
    \mathrm{Conf}(c,\,t) \geq 1
    \;\land\;
    \forall p \in \mathcal{P}^H : p(c,\,t) = 0
  \right].
\]
$\mathrm{Accept}(c,t) = 1$ is the condition under which HSP-Agile declares
the synthesis attempt successful.
\end{definition}

\paragraph{Example.}
A repaired \code{classify\_signal} with $\mathrm{Conf}=1$ still fails
acceptance if RF07 (\emph{unconditional success}) fires on a hard-coded
\texttt{return "Normal"}.
Conversely, a candidate that clears the pattern screen but fails one witness
also fails $\mathrm{Accept}$.
This is the source of the strict-success paradox in Chapter~\ref{ch:results}:
mean conformance can be high while acceptance remains selective.

The conjunction of formal conformance and pattern safety reflects two error
modes: (i)~functional errors on specific inputs, caught by witnesses; and
(ii)~structural anti-patterns, caught by static screening.
Neither check alone is sufficient.

Those pieces assemble into the main computational problem: find an acceptable
candidate within a fixed LLM budget, or return the best-effort candidate when
the budget is exhausted.

\begin{definition}[Bounded Acceptance Problem]
\label{def:bap}
Given:
\begin{itemize}
  \item a task $t = (\mathcal{S}_t, \mathcal{X}_t, \mathcal{Y}_t)$;
  \item a budget $K \in \mathbb{N}^+$ bounding the number of LLM calls; and
  \item a synthesis oracle $\mathrm{LLM}(\cdot)$ that, given a task and a
        feedback object, generates a candidate implementation;
\end{itemize}
the \emph{Bounded Acceptance Problem} (BAP) is to find a candidate
\[
  c^* = \mathrm{LLM}\!\bigl(t,\;\mathrm{feedback}_1, \ldots,
                              \mathrm{feedback}_{k-1}\bigr)
\]
with at most $K$ calls to $\mathrm{LLM}$ such that $\mathrm{Accept}(c^*, t) = 1$.
If no such candidate exists within budget, return the best-effort candidate
\[
  c^* \;=\; \arg\max_{c \,\in\, C_K} \mathrm{Conf}(c,\,t),
\]
where $C_K = \{c_1, \ldots, c_K\}$ is the set of all candidates generated
during the $K$ attempts.
\end{definition}

The feedback object $\mathrm{feedback}_k$ passed to the LLM on attempt $k+1$
is a structured JSON record containing: (i) the list of counterexamples from
attempt $k$; (ii) the list of detected pattern violations; and (iii) the
attempt index $k$.  The structure and content of feedback messages are
described in detail in Chapter~\ref{ch:method}.

The pipeline is therefore an instance of counterexample-guided inductive
synthesis (CEGIS) \citep{solarlezama2008sketching, lynn2024verifierloop}, where the verifier is the
SMT-generated case set and the pattern checker, and the synthesiser is the
language model.
Unlike classical CEGIS, however, the synthesiser is stochastic: the same prompt
may produce different outputs on successive calls, with no guarantee of
convergence within budget $K$.
The BAP makes this limitation explicit by specifying a graceful fallback when
the budget is exhausted.

% ---------------------------------------------------------------------------
\section{Prevention and Quality Metrics}
\label{sec:form:metrics}
% ---------------------------------------------------------------------------

Generation success is the wrong primary metric for a prevention system.
What matters is whether faulty candidates are kept out of release, and whether
correct candidates are not rejected by accident.
Those two rates are measured over a mutant population
\citep{demillo1978mutation, ammann2016testing} and reported in
Chapter~\ref{ch:experiments}.

\begin{definition}[Mutant Population]
\label{def:mutants}
A \emph{mutant population} $\mathcal{M}$ is a finite set of pairs
$(c_m, t_m)$, each consisting of a candidate and a task, where at least
one of the two has been artificially perturbed.
Two sub-populations matter:
\begin{itemize}
  \item \emph{Specification-confusion mutants} $\mathcal{M}^S$: the task
        $t_m$ is a mutated version of a reference task, obtained by
        modifying scenario guards or output bindings in $\mathcal{S}_t$.
        The candidate $c_m$ is a correct implementation of the
        \emph{original} task, so it is faulty with respect to $t_m$.
  \item \emph{Implementation-screening mutants} $\mathcal{M}^I$: the
        task $t_m$ is unchanged, but $c_m$ is a deliberately mutated
        version of a reference implementation, obtained by applying a
        standard mutation operator (e.g., guard-order swap,
        boundary-off-by-one).
\end{itemize}
A mutant pair $(c_m, t_m)$ is \emph{genuinely faulty}, written
$\mathrm{Faulty}(m) = 1$, if the reference specification or the reference
implementation has been perturbed.
\end{definition}

PDR asks a recall-like question: of all mutants, how many does the pipeline
correctly refuse to accept?

\begin{definition}[Prevention Detection Rate]
\label{def:pdr}
The \emph{Prevention Detection Rate} (PDR) is the fraction of mutants that
the pipeline correctly rejects:
\[
  \mathrm{PDR}
  \;=\;
  \frac{\bigl|\bigl\{m \in \mathcal{M} : \neg\,\mathrm{Accept}(c_m,\,t_m)\bigr\}\bigr|}
       {|\mathcal{M}|}.
\]
In a sprint review, high PDR means fewer faulty candidates reach release.
\end{definition}

FAR asks the complementary precision-like question: among genuinely faulty
mutants, how many still slip through $\mathrm{Accept}$?

\begin{definition}[False Accept Rate]
\label{def:far}
The \emph{False Accept Rate} (FAR) is the fraction of genuinely faulty
mutants that the pipeline incorrectly accepts:
\[
  \mathrm{FAR}
  \;=\;
  \frac{\bigl|\bigl\{m \in \mathcal{M} :
      \mathrm{Accept}(c_m,\,t_m) = 1
      \;\land\; \mathrm{Faulty}(m) = 1
    \bigr\}\bigr|}
       {|\mathcal{M}|}.
\]
In a sprint review, low FAR means the gate is not rubber-stamping mutants.
\end{definition}

PDR and FAR are complementary but not redundant \citep{okun2007mutation}.
PDR counts every rejected mutant, including trivial crashes; FAR focuses on
faulty mutants that still receive $\mathrm{Accept}=1$.

A related metric quantifies the contribution of the formal case-based check
alone, independent of the pattern screening layer.

\begin{definition}[Mutation Kill Rate]
\label{def:mkr}
The \emph{mutation kill rate} is the fraction of mutants detected by formal
conformance checking:
\[
  \mu_k
  \;=\;
  \frac{\bigl|\bigl\{m \in \mathcal{M} : \mathrm{Conf}(c_m,\,t_m) < 1\bigr\}\bigr|}
       {|\mathcal{M}|}.
\]
\end{definition}

The distinction between $\mu_k$ and PDR is important.  $\mu_k$ measures
whether the formal case-based check alone is sufficient to detect a given
mutant (i.e., at least one generated case produces a wrong output).  PDR
measures whether the \emph{full pipeline} — including pattern screening —
correctly rejects the candidate.  A mutant with $\mathrm{Conf}(c_m, t_m) = 1$
but a detected pattern violation contributes to PDR but not to $\mu_k$,
because the conformance check alone would have passed it.  The gap
$\mathrm{PDR} - \mu_k$ therefore quantifies the additive value of the
pattern-screening layer over pure case-based testing.

% ---------------------------------------------------------------------------
\section{Benchmark Formalization}
\label{sec:form:benchmark}
% ---------------------------------------------------------------------------

The evaluation uses a synthetic benchmark of hard tasks built to stress FSF
precedence sensitivity \citep{li2023iet}.
The construction below makes that stress precise and proves that correct
solutions must respect scenario ordering.

\begin{definition}[Precedence-Sensitive Task]
\label{def:prec-sensitive}
A task $t$ is \emph{precedence-sensitive} if there exist two scenarios
$s_i, s_j$ with $i < j$ such that their guard regions overlap:
\[
  \bigl\{x \in \mathcal{X}_t : g_i(x) = 1\bigr\}
  \;\cap\;
  \bigl\{x \in \mathcal{X}_t : g_j(x) = 1\bigr\}
  \;\neq\; \emptyset.
\]
\end{definition}

In a precedence-sensitive task, the output on inputs in the overlapping region
depends entirely on which guard is evaluated first.  An implementation that
checks $g_j$ before $g_i$ will, on those inputs, return $\phi_j(x)$ rather
than the correct $\phi_i(x)$.  This is precisely the class of error that
naive LLM synthesis frequently introduces: the model may faithfully reproduce
all guard conditions and output mappings from the specification text, while
inadvertently evaluating them in a different order.

\begin{proposition}
\label{prop:order-required}
For any precedence-sensitive task $t^H$, a strictly correct implementation
must evaluate the scenario guards in the exact order prescribed by
$\mathcal{S}_{t^H}$.
\end{proposition}

\begin{proof}
Let $x^*$ be any input in the overlapping region of $s_i$ and $s_j$
($i < j$).  By Definition~\ref{def:oracle}, the oracle maps
$g_t(x^*) = \phi_i(x^*)$.  Suppose an implementation $c$ evaluates $g_j$
before $g_i$.  Since $g_j(x^*) = 1$, the implementation returns
$\phi_j(x^*)$.  Because the task is precedence-sensitive (the two scenarios
cover an overlapping region of the input space), the specification is
non-trivially ordered, and it need not be the case that $\phi_i(x^*) =
\phi_j(x^*)$ — indeed, if they were always equal on the overlap, the overlap
would be semantically irrelevant and the task would not test ordering at all.
Hence $f_c(x^*) \neq g_t(x^*)$, and $c$ does not strictly succeed.
\end{proof}

Hard synthetic tasks are constructed as follows.
Each task $t^H$ has the following structure:

\begin{itemize}
  \item \textbf{Input variables}: five non-negative integers
        $a, b, c, d, e \in \mathbb{N}$.
  \item \textbf{Output variables}: three integer quantities —
        $\mathit{risk} \in \{0,1,2\}$,
        $\mathit{action} \in \{0,1,2\}$, and
        $\mathit{score} \in [0, 100]$ — giving $\mathcal{Y}_t \subseteq
        \{0,1,2\}^2 \times [0,100]$.
  \item \textbf{Scenario count}: up to $n = 9$ ordered scenarios.
  \item \textbf{Threshold parameters}: seven threshold values
        $\theta_1, \ldots, \theta_7$ sampled uniformly at random from
        $[1, 100] \cap \mathbb{Z}$, used to construct the guard predicates
        (e.g., $a > \theta_1$, $b \leq \theta_2 \land c > \theta_3$).
  \item \textbf{Feasibility guarantee}: all scenarios are satisfiable, verified
        by a Z3 feasibility check during task generation.  Any candidate
        $\theta$ vector that renders a scenario unsatisfiable is discarded
        and resampled.
  \item \textbf{Consistency validation}: the generated task passes a
        scenario-consistency validator that confirms each generated case
        belongs to exactly one scenario and that the ``others'' guard
        correctly covers all remaining inputs.
\end{itemize}

By Proposition~\ref{prop:order-required}, every hard task in the benchmark
requires an implementation to evaluate guards in the specified order.  An
LLM that does not attend to the ordering of scenarios in the prompt will
therefore produce incorrect implementations, which are caught by the
SMT-generated test suite.

% ---------------------------------------------------------------------------
\section{Computational Considerations}
\label{sec:form:complexity}
% ---------------------------------------------------------------------------

Formal checking must not erase the prevention benefit by dominating runtime.
The cost accounting below confirms that Z3 and pattern screening remain cheap
relative to LLM inference \citep{demoura2008z3, barrett2018smt}.

\paragraph{Case Generation.}
For each of the $n$ scenarios, the Z3 solver is invoked with a linear
arithmetic formula over $d$ integer variables.  Linear integer arithmetic
(LIA) is decidable, and Z3's LIA solver runs in time that is polynomial in
the formula size for the types of constraints that arise from threshold-based
FSF guards.  In practice, with $n \leq 9$ and $d = 5$, each call to Z3
completes in under $0.1$ seconds, giving a total case generation time of
$O(n \cdot d^2)$ solver steps in the worst case, or approximately
$0.5$--$1$ seconds per task.

\paragraph{Conformance Checking.}
Given the case set $\mathcal{D}(t)$ of size $n$, checking conformance
requires executing the candidate $c$ on each input $x_i$ and comparing the
output to $g_t(x_i)$.  This is a simple lookup-and-compare operation with
cost $O(n)$ function evaluations and $O(n)$ comparisons.  For $n \leq 9$,
this step is effectively instantaneous.

\paragraph{Pattern Matching.}
The pattern screening layer applies $|\mathcal{P}^H| = 14$ pattern checkers
to the source code of candidate $c$.  Each pattern is encoded as a regular
expression or a simple abstract-syntax-tree traversal.  The cost is $O(L
\cdot P)$ where $L$ is the length of the candidate source code and $P =
|\mathcal{P}^H|$.  For typical LLM-generated functions of 30--80 lines,
this step completes in well under $0.01$ seconds.

\paragraph{Total Pipeline Cost.}
The dominant cost at every iteration is the LLM API call, which takes
approximately $10$ seconds per attempt in the reported experiments.
Z3 solving and conformance checking add less than $1$ second combined;
pattern matching is negligible.
With budget $K = 3$, expected runtime is about $30$ seconds per task, of
which more than $97\%$ is LLM inference.
Formal checking is therefore not the bottleneck: larger case sets or richer
pattern libraries would not meaningfully change throughput.

\paragraph{Summary.}
The cost structure can be written as
\[
  T_{\mathrm{pipeline}}
  \;\approx\;
  K \cdot T_{\mathrm{LLM}}
  \;+\; O(n \cdot d^2) \cdot T_{\mathrm{Z3}}
  \;+\; O(n) \cdot T_{\mathrm{exec}}
  \;+\; O(L \cdot P) \cdot T_{\mathrm{pattern}},
\]
where $T_{\mathrm{LLM}} \gg T_{\mathrm{Z3}} \gg T_{\mathrm{exec}} \approx
T_{\mathrm{pattern}}$, confirming that formal checking adds less overhead
than a single LLM call.

% ---------------------------------------------------------------------------
\section*{Chapter Summary}
% ---------------------------------------------------------------------------

The formal picture is now complete enough to build on.
A task is an ordered scenario specification over typed domains; SpecIR is the
shared hinge across notations; first-match semantics yield an executable
oracle; SMT witnesses populate a finite case set with full scenario coverage
(Lemma~\ref{lem:coverage}); and release is the conjunctive predicate
$\mathrm{Accept}$ under a bounded search
(Definition~\ref{def:bap}).
PDR, FAR, and mutation kill rate measure prevention rather than generation
success.
Every hard benchmark task requires strict scenario ordering
(Proposition~\ref{prop:order-required}), while formal checking remains cheap
relative to LLM inference.
What remains is operational: how an LLM participates in that loop, how
failures become Semantic Feedback IR, and how the four stages assemble into a
method.
