%% ch06_experimental_setup.tex  —  Chapter 6: Experimental Setup
%% HSP-Agile CCF-B Report, East China Normal University, 2026
%% Ye Xujun

\chapter{Experimental Setup}
\label{ch:experiments}

This chapter is the measurement protocol for the three Tier-1 claims.
It fixes what is compared and how prevention is scored so
Chapter~\ref{ch:results} can answer \emph{why we should believe the ideas},
not merely that a system runs.
Hardware, modes, metrics, and E1--E9 tracks are stated once; threats are
previewed here and analysed in Chapter~\ref{ch:discussion}.

\begin{inlinetable}{Claim-to-experiment map (Tier-1 focus).}
  \label{tab:claim-map}
  \footnotesize
  \begin{tabular}{llp{0.42\linewidth}}
    \toprule
    Claim & Experiments & What would falsify it \\
    \midrule
    C1 (semantic regions) & E3, E4, E8 & Gap vanishes under dense overlap;
      adapters fail to transfer \\
    C2 (typed feedback) & E5, E6, A3 & Full IR $\not>$ test-only; repair
      unused \\
    C3 (conjunctive gate) & E1 Strict, E2 PDR/FAR, A2 & High Conf.\ with
      high FAR; guard adds nothing \\
    \bottomrule
  \end{tabular}
\end{inlinetable}

%=============================================================================
\section{Research Questions}
\label{sec:exp:rq}
%=============================================================================

Each RQ answers a reviewer concern and maps to the claims above.

\paragraph{RQ1 (Overall Effectiveness).}
Does the three-insight loop improve formal conformance over one-shot and
test-feedback baselines? (E1, E2; supports C1--C3 jointly.)

\paragraph{RQ2 (Fault Class Prevention).}
Which defect classes does the conjunctive gate catch, and which remain?
(E2, E7, E9; primarily C3, residual analysis for C1/C2.)

\paragraph{RQ3 (Component Attribution).}
How much do typed feedback and the formal checker contribute independently?
(E5, E6, A1--A3; primarily C2, with C3 via A2/PDR.)

\paragraph{RQ4 (Complexity Scaling).}
Does the advantage grow where semantic regions are hardest to sample?
(E3, E4; primarily C1.)

\paragraph{RQ5 (Quality--Overhead Trade-off).}
Is the latency cost justified by the conformance gain?
(E1 Pareto; practical bound on all three claims.)

%=============================================================================
\section{Hardware and Software Environment}
\label{sec:exp:environment}
%=============================================================================

Prevention claims require fixed oracle conditions: the same solver, the same
timeouts, and the same parallel schedule for every mode.
All runs used a Windows~10 (build~26220) desktop with a 6-core processor,
Python~3.12, and one \code{multiprocessing} worker per core.
Z3~4.16~\citep{demoura2008z3,barrett2018smt} was invoked through
\code{z3-solver} with a 10-second per-call timeout; an \code{unknown} result
is treated as non-constraining rather than as a hard failure.

%=============================================================================
\section{Language Model Configuration and Run Protocol}
\label{sec:exp:protocol}
%=============================================================================

\subsection{Language Model Configuration}

All evaluated LLM-based modes (B1, B2, M, A1, A2, A3) used
Qwen2.5-27B-Instruct~\citep{yang2024qwen25} (\textsf{Qwen-27B}) accessed through an
OpenAI-compatible LLM gateway.
Temperature was set to $T_\text{gen} = 0.7$ for all generation calls and to
$T_\text{repair} = 0.0$ for all repair calls, following the prompting strategy
described in Chapter~\ref{ch:method}.
The maximum token budget per LLM call was 4096 tokens.
The budget parameter $K = 3$ was fixed for all LLM-based modes: in mode~M, the repair
loop may invoke the LLM up to three times per task, consuming between 1 and 3 API
calls depending on when the loop terminates.

\subsection{Experiment Identifier and Scale}

The primary experiment was conducted under the run identifier
\code{run\_hard\_full\_parallel\_v1}.
In total, 1{,}200 runs were executed: 10 modes $\times$ 120 tasks, each run
processed independently by a dedicated worker process.
The core seven-mode matrix (\code{run\_hard\_full\_parallel\_v1}) covers
B0--B2, M, and A1--A3; extended baselines B3--B5 were evaluated under
\code{run\_ccf\_b\_extended\_v1} (repeat~0 merged into the reported tables
after \texttt{config\_for\_mode} was corrected for those modes).
Run artefacts (prompts, LLM responses, Z3 query logs, evaluation traces, timing
records) were serialised to disk in JSONL format for
reproducibility~\citep{ammann2016testing}.
All random-number generation used fixed seed~42 for task generation and prompt
construction; the LLM API itself is not seeded, but temperature~0.0 for repair
calls makes those responses near-deterministic in practice.

Multi-seed stability (E12) and uniform random-benchmark (E10) protocols are
described in Sections~\ref{sec:exp:e10} and~\ref{sec:exp:e12}.

\subsection{Reproducibility Measures}

Determinism was enforced at three levels: fixed-seed task generation, deterministic
evaluation harness (Z3 case generation, mutation testing), and non-parametric
statistical tests robust to residual LLM non-determinism.
The \code{attempt\_history} field in every result JSONL records per-iteration
conformance and counterexample counts, enabling E5 repair dynamics analysis without
re-running experiments.
Data preparation, aggregation, and plotting each run as a distinct pipeline
stage (\code{prepare\_paper\_data.py}, \code{prepare\_mechanism\_data.py},
\code{plot\_mpl\_figures.py}).
All reported tables and figures are generated from the canonical
\code{run\_hard\_full\_parallel\_v1} artifact (120 tasks $\times$ 7 modes),
\code{run\_ccf\_b\_extended\_v1} (B3--B5, repeat~0), and
\code{prevention\_full\_v1} prevention evaluation.
Long-running mechanism experiments (E6 feedback variants, E8 generalisation) write
incremental \code{results.jsonl} and a live \code{progress.json} with completed/total
counts and ETA; use \code{python -u} for unbuffered console output.
Full reproduction instructions are provided in Appendix~\ref{app:reproducibility}.

%=============================================================================
\section{Benchmark Construction}
\label{sec:exp:benchmark}
%=============================================================================

\subsection{Design Rationale}

Ordered FSF artefacts are rare in public repos, so the corpus is constructed.
Three requirements: discriminative (baselines fail non-trivially), formally
valid (Z3-satisfiable), and diverse (varied orderings and boundaries).
DafnyBench~\citep{loughman2024dafnybench} targets annotation-heavy
verification, not FSF precedence---hence a domain-specific set.

120 hard tasks (from 180 candidates) survive a Z3 filter that keeps only
non-empty overlap regions: five inputs, three outputs, eight ordered
scenarios, annotated by overlap density, guard-complexity label, and
scenario count.
B0 at 27.5\% strict success confirms genuine difficulty.

\subsection{Base Tasks and Hard Synthetic Generation}

Construction began with 20 base SOFL/FSF tasks extracted by manual inspection from published
example specifications~\citep{liu2004soflbook,li2022qrs_companion}.
Each base task defines a small module with 3--5 inputs, 2--3 outputs, and 5--8 FSF
scenarios.
The base tasks were retained as a development set but excluded from primary evaluation
to avoid information leakage.

The \emph{HardTaskGenerator} produces synthetic FSF tasks with exactly 5 integer
inputs $(a, b, c, d, e) \in [0, 100]^5$, 3 outputs
(\textit{risk}, \textit{action}, \textit{score}), and exactly 8 ordered FSF scenarios.
Difficulty is modulated by two design choices.
\emph{Precedence-sensitive guard regions:} each scenario's pre-condition guard overlaps
with at least one other guard, so correct evaluation requires strict priority
ordering---directly targeting scenario-order violations (SOV) from
Chapter~\ref{ch:formalization}.
\emph{Z3 satisfiability filter:} before admission, Z3 verifies that all 8 scenarios are
individually satisfiable; tasks failing this check are discarded.

Figure~\ref{fig:benchmark-pipeline} is the construction path from sampled
guards through Z3 filtering to the final 120-task set.
The filter retains only overlap-heavy, satisfiable tasks---the regime where
ordering defects actually occur.

\begin{figure}[t]
  \centering
  \tikzinput{diagrams/tikz/benchmark_pipeline}
  \caption{\textbf{Key observation:} the Z3 filter keeps only overlap-heavy,
    satisfiable tasks---the regime where ordering defects actually occur.
    \textbf{How to read:} left-to-right from sampled guards through filtering
    to the final 120-task set.}
  \label{fig:benchmark-pipeline}
\end{figure}

The generator produced 180 candidate tasks, of which 60 were discarded by the Z3
satisfiability filter, leaving the final benchmark of 120 hard synthetic tasks.

\subsection{Random Benchmark (E10)}
\label{sec:exp:e10}

To address overlap-filter bias, a \emph{uniform random} benchmark of 100 tasks
was generated from the same \emph{HardTaskGenerator} family but \emph{without}
the Z3 overlap-density filter (\texttt{--no-filter} in
\texttt{build\_benchmark.py}).
Tasks were annotated with overlap rate, boundary density, and scenario count for
post-hoc stratification.
Evaluation runs B1, B2, and M under the same $K{=}3$, model, and token budget
as E1 (\texttt{run\_e10\_random\_v1} on
\texttt{benchmarks/random\_tasks\_annotated.json}).
The decision question is whether M's advantage over B2 persists on uniformly
sampled tasks or concentrates in the overlap-intensive tertiles already
observed in E3.

\subsection{External SOFL Benchmark (E11)}
\label{sec:exp:e11}

To bound author-construction bias, we curated $n{=}22$ hand-authored FSF tasks
in \texttt{benchmarks/external\_sofl.json} from published Agile-SOFL and IET
fault-prevention patterns~\citep{liu2004soflbook,li2022qrs_companion,li2023iet_faultprevention}.
Each task records \texttt{externalProvenance} (source citation, book section);
none are generated by \texttt{HardTaskGenerator}.
Task IDs are disjoint from the 120-task E1 hard set.
A parallel export \texttt{benchmarks/manual\_heldout.json} (same $n{=}22$, zero E1
overlap) is reserved for vendor-example provenance when the Agile-SOFL submodule is
available; in this artifact package both files reference the curated external corpus
because \texttt{vendor/agile-sofl-toolchain/examples} is not bundled.
Evaluation (\texttt{run\_e11\_external\_v1}) runs B1, B2, and M under the same
$K{=}3$ budget as E1.

\subsection{Generalisation Subset (E8)}

\paragraph{Real-derived benchmark subset.}
External validity is further probed on 40 tasks derived from publicly available
coding benchmarks: 20 from HumanEval~\citep{chen2021codex} and 20 from
MBPP~\citep{austin2021mbpp}.
We selected problems using the following protocol: (i)~the reference solution
contains 3--6 ordered conditional branches; (ii)~each branch maps naturally to an
FSF scenario with a guard predicate involving linear inequalities or equalities;
(iii)~no external library imports or recursive depth~$>2$.
Reference implementations were manually converted to FSF format and cross-checked
against the original test suites.
The conversion script (\texttt{scripts/convert\_humaneval\_mbpp.py}) and
the selection protocol (\texttt{artifacts/BENCHMARK\_SELECTION\_PROTOCOL.md})
are released with the artifact package.

\subsection{Illustrative Task Example}

\code{classify\_signal} was the two-input monitoring alert; the hard benchmark
is the same first-match skeleton scaled up.
Five inputs $(a,b,c,d,e)$, three outputs, eight nested guards---still
``most-specific fires first,'' still the overlap region where ordering bugs
hide.
The illustrative task below is a financial risk classifier
(\code{a}--\code{e} $\in [0,100]$; outputs \code{risk}, \code{action},
\code{score}), ordered most-specific to least-specific:

\begin{enumerate}
  \item High credit AND low debt AND low volatility $\Rightarrow$ (risk=0, action=0, score=90)
  \item High credit AND low debt $\Rightarrow$ (risk=0, action=0, score=75)
  \item High credit AND moderate debt $\Rightarrow$ (risk=1, action=1, score=60)
  \item Low credit AND high debt AND high volatility $\Rightarrow$ (risk=3, action=2, score=15)
  \item Low credit AND high debt $\Rightarrow$ (risk=2, action=2, score=30)
  \item Low credit $\Rightarrow$ (risk=2, action=1, score=40)
  \item Moderate credit AND high exposure $\Rightarrow$ (risk=1, action=1, score=55)
  \item (default) $\Rightarrow$ (risk=1, action=0, score=65)
\end{enumerate}

Scenario~1 and Scenario~2 have overlapping guards.
An implementation evaluating guards in reverse order produces \code{score}=75 instead
of 90 on Scenario~1 inputs---a scenario-order violation detectable by the pattern
guard and by Z3 witnesses in the overlap region
(e.g., \code{a=85, b=15, c=10, d=80, e=20}).
A one-shot LLM may handle 7 of 8 scenarios correctly yet fail on this overlap
because it did not fully parse priority ordering from the specification
text~\citep{chen2021codex,dakhel2023copilot}; the CEGIS repair loop can correct this
in a single iteration given a Z3 boundary input~\citep{solarlezama2008sketching,sun2024clover}.

\subsection{Benchmark Characteristics}

Table~\ref{tab:benchmark-stats} summarises the 120-task benchmark;
the mean guard region overlap rate is 1.8 scenarios per input point.

\begin{table}[H]
  \caption{Summary statistics of the 120-task hard benchmark.}
  \label{tab:benchmark-stats}
  \centering
  \thesistable{%
    \begin{tabular}{lc}
      \toprule
      Property & Value \\
      \midrule
      Total tasks & 120 \\
      Inputs per task & 5 \\
      Outputs per task & 3 \\
      Scenarios per task & 8 \\
      Mean guard-overlap rate (scenarios/point) & 1.8 \\
      Tasks rejected by Z3 filter & 60 \\
      Tasks accepted & 120 \\
      Conformance cases per task (\code{strict\_eval}) & 64 \\
      \bottomrule
    \end{tabular}%
  }
\end{table}

%=============================================================================
\section{Compared Methods}
\label{sec:exp:methods}
%=============================================================================

The evaluation question is which feedback source actually repairs precedence
faults: none, unit tests, self-critique, execution traces, or
specification-indexed SMT witnesses.
Ten modes are evaluated in E1 (B0--B5, M, A1--A3).
The design space spans formal reference (B0), LLM baselines (B1--B5), full
HSP-Agile (M), and three ablations (A1--A3).

\begin{table}[H]
  \caption{Compared modes.
    Columns indicate whether each component is active.
    B3--B5 are evaluated in E1-ext (\code{run\_ccf\_b\_extended\_v1},
    repeat~0 merged into Table~\ref{tab:main-results}).}
  \label{tab:modes}
  \centering
  \thesistable{%
    \footnotesize
    \renewcommand{\arraystretch}{1.15}
    \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l l c c c@{}}
      \toprule
      Mode & Description & Formal checker & Pattern guard & Repair loop \\
      \midrule
      B0 & Reference (template) & eval only & --- & --- \\
      B1 & One-shot LLM & --- & --- & --- \\
      B2 & Test-feedback & --- & --- & \checkmark (tests) \\
      B3 & Self-Refine~\citep{madaan2023selfrefine} & --- & --- & \checkmark (self-critique) \\
      B4 & Self-Debug~\citep{chen2023selfdebug} & --- & --- & \checkmark (execution trace) \\
      B5 & Reflexion-lite~\citep{shinn2023reflexion} & --- & --- & \checkmark (verbal memory) \\
      B6 & VerifierLoop-FSF (REV-7) & \checkmark & --- & \checkmark (witness cex) \\
      M  & Full \textsc{SgDP}/HSP-Agile & \checkmark & \checkmark & \checkmark (IR) \\
      A1 & Ablation: no formal checker & --- & \checkmark & \checkmark (IR) \\
      A2 & Ablation: no pattern guard & \checkmark & --- & \checkmark (IR) \\
      A3 & Ablation: no repair loop & \checkmark & \checkmark & --- \\
      \bottomrule
    \end{tabular*}%
  }
\end{table}

\paragraph{B0 --- Reference Implementation.}
B0 provides a formal upper-bound reference.
For each task, a deterministic template renderer encodes each FSF scenario as a guarded
branch in priority order; no LLM or iterative loop is involved.
The formal checker is applied \emph{ex post} to compute $\mathit{Conf}(\cdot)$.

\paragraph{B1 --- One-Shot LLM.}
B1 makes a single LLM API call with the full FSF specification, parses the response,
and passes it directly to the evaluator with no feedback loop.

\paragraph{B2 --- LLM with Test-Feedback Loop.}
B2 extends B1 with an iterative test-feedback loop~\citep{zhu1997specification}.
After each generation, unit tests derived from FSF pre/post-conditions are executed;
failures are appended to the prompt for the next LLM call until all tests pass or
$K=3$ is exhausted.
B2 uses no Z3 and no pattern guard.

\paragraph{B3 --- Self-Refine Baseline.}
B3 (Self-Refine style~\citep{madaan2023selfrefine}): after each synthesis attempt, the
same LLM is asked to critique its output and revise without any external checker.
The critique prompt requests a structured self-assessment but provides no Z3
counterexamples, pattern-guard signals, or specification-level annotation.

\paragraph{B4 --- Self-Debug Baseline.}
B4 (Self-Debug style~\citep{chen2023selfdebug}): the candidate is executed on B0
reference test cases and execution failures are fed back as natural-language without
specification-level annotation.
Unlike B2, B4 uses only reference-implementation observable behaviour rather than
FSF-derived tests.
All three extended baselines B3--B5 use the same prompt template family as M,
the same budget~$K$, and the same per-call token cap; results appear in
Table~\ref{tab:main-results} (repeat~0, \code{run\_ccf\_b\_extended\_v1}).

\paragraph{B5 --- Reflexion-Lite Baseline.}
B5 (Reflexion-style~\citep{shinn2023reflexion}): a lightweight verbal memory
buffer records prior attempt summaries across repair iterations without Z3 or
pattern-guard signals.
Like B3 and B4, B5 isolates whether unstructured self-improvement can match
specification-indexed Semantic Feedback IR under equal budget.

\paragraph{B6 --- VerifierLoop-FSF Baseline (REV-7).}
B6 adapts verifier-in-the-loop repair to FSF~\citep{sun2024clover}:
the formal checker runs during the repair loop with the same witness budget as M
(\texttt{formal\_max\_cases}${=}24$), but repair prompts expose only structured
SMT counterexamples (scenario guard, inputs, expected outputs) via the
\texttt{verifier\_loop} feedback variant---without Semantic Feedback IR fields
(\texttt{violation\_type}, \texttt{reason}, \texttt{suggested\_fix}) and without
the pattern guard.
Implementation: \texttt{src/baselines/verifier\_loop\_fsf.py};
evaluation: \texttt{run\_b6\_full\_v1} ($n{=}100$ hard tasks) and
\texttt{run\_b6\_stratified\_v1} (30-task E12 subset vs.\ B2/M).

\paragraph{M --- Full HSP-Agile Pipeline.}
Mode~M runs the complete pipeline (Chapters~\ref{ch:method} and~\ref{ch:implementation}):
LLM generation, Z3 counterexample inputs, pattern-guard screening, and CEGIS repair
with Semantic Feedback IR until all checks pass or $K=3$ is reached.

\paragraph{A1 --- Ablation: No Formal Check.}
A1 replaces Z3-based case generation with the B2 unit-test harness; pattern guard and
repair loop remain active, isolating Z3-guided feedback.

\paragraph{A2 --- Ablation: No Pattern Guard.}
A2 disables the pattern guard; candidates are evaluated by Z3 and passed directly to
the repair loop, isolating structural screening.

\paragraph{A3 --- Ablation: No Repair Loop.}
A3 calls the LLM exactly once (as B1) but applies Z3 and the pattern guard without
sending corrective feedback, isolating the CEGIS repair loop.

%=============================================================================
\section{Mutant Design}
\label{sec:exp:mutants}
%=============================================================================

Prevention metrics need faults that look locally plausible yet violate global
precedence---not random syntax noise.
Two tracks supply them.
\emph{Specification-confusion} mutants perturb the task: SNO swaps or negates
scenario order, ORO reverses outputs, MCO drops a condition, BCO offsets a
boundary.
\emph{Implementation-screening} mutants perturb the candidate: ICO flips a
condition, WRO reorders returns, MBO drops a branch, DRO reverses a
dependency.
Each mutant applies one first-order operator to a known-correct artefact;
non-compilable mutants are excluded and reported separately.

\paragraph{Miniature example (how PDR/FAR are counted).}
Start from a correct \code{classify\_signal} implementation that checks
$\mathit{level}<0$ first.
An \emph{SNO} specification-confusion mutant swaps $S_1$ and $S_2$ in the
oracle text; the pipeline's $\mathrm{Accept}$ gate rejects the candidate that
still implements the original order, counting as a true prevention (PDR
numerator).
An \emph{WRO} implementation-screening mutant reorders the candidate's
\texttt{if} branches while leaving the specification untouched; again
$\mathrm{Accept}$ rejects it.
A false acceptance (FAR numerator) would occur only if a faulty mutant still
passed every witness and the pattern screen---precisely what Theorem~\ref{thm:far-bound}
bounds.

%=============================================================================
\section{Evaluation Metrics}
\label{sec:exp:metrics}
%=============================================================================

\subsection{Primary Metrics}

Different metrics answer different questions.
\emph{Conformance} measures continuous quality on witnesses;
\emph{strict success} measures the conjunctive release gate;
\emph{PDR/FAR} measure prevention on mutants;
latency and LLM calls measure cost.

\paragraph{Strict Formal Conformance.}
The formal conformance $\mathit{Conf}(c, t) \in [0,1]$
(Definition~\ref{def:conformance}) is the fraction of strict evaluation cases that
candidate~$c$ satisfies on task~$t$; mean $\overline{\mathit{Conf}}$ is reported
across all 120 tasks.
This is the primary discriminating metric on the hard benchmark.

\paragraph{Strict Success Rate.}
For a task~$t$ and mode~$m$, the run is a \emph{strict success} if the returned
candidate passes every one of the 64 Z3-generated evaluation cases
(\code{strict\_eval} set).
The strict success rate is the fraction of 120 tasks meeting this criterion;
modes that enable the pattern screen additionally require a clear
high-severity screen for $\mathrm{Accept}=1$ (Equation~\eqref{eq:accept}).

\paragraph{Mutation Kill Rate.}
Given candidate~$c$ for task~$t$, $n_\mu$ syntactic mutants are injected and
checked for whether the formal checker classifies each as non-conforming.
The mutation kill rate $\mu_k$ is the fraction killed; it measures checker precision,
not generator quality.

\paragraph{LLM Call Count and Latency.}
LLM call count is the mean API calls per task (1.0 for B1; 1--3 for repair modes).
Latency is wall-clock time from task start to final verdict, including LLM, Z3, pattern
guard, and orchestration overhead.

\paragraph{PDR and FAR (Prevention Metrics).}
PDR is the fraction of faulty mutants for which
$\neg\mathrm{Accept}(c_m,t_m)$ (correctly blocked).
FAR is the fraction of faulty mutants for which $\mathrm{Accept}(c_m,t_m)$
still holds (escaped into release); see Definitions~\ref{def:pdr}--\ref{def:far}
and the miniature example above.
Both are reported per mutant operator in E2 and aggregated in E1.

%=============================================================================
\section{Evaluation Protocol}
\label{sec:exp:eval-protocol}
%=============================================================================

Each experiment answers one decision question; artifact paths live in
Appendix~\ref{app:reproducibility}.

\paragraph{E1 --- Overall Effectiveness.}
Does the full loop beat one-shot and test-feedback on Conf, PDR, FAR, and
latency?  120 tasks $\times$ 10 modes (1{,}200 runs: core matrix plus B3--B5
from \code{run\_ccf\_b\_extended\_v1}); $K=3$, $T_\text{gen}=0.7$,
$T_\text{repair}=0.0$.

\paragraph{E2 --- Full Prevention Evaluation.}
Which mutant operators does each mode catch?  120 tasks $\times$ 6 modes
$\times$ 8 operators, reporting PDR/FAR per operator.

\paragraph{E3 --- Specification Complexity.}
Does the M--baseline gap grow with overlap?  Conf stratified by density
tertile and guard-complexity label.

\paragraph{E4 --- Boundary Density.}
Is SMT sampling necessary, or would denser random tests suffice?
Coverage of SMT vs.\ random across budgets $\{4,8,16,32,64\}$ and three
density tiers.

\paragraph{E5 --- Repair Dynamics.}
Where does budget $K=3$ pay off?  $\mathrm{Conf}(k)$ for M vs.\ B2 at
$k=1,2,3$ from logged \code{attempt\_history}.

\paragraph{E6 --- Feedback Variant Comparison.}
Is the gain from structure or from longer prompts?  120 tasks $\times$ three
variants under mode~M: A (test only), B (test + expected), C (full Semantic
Feedback IR).

\paragraph{E7 --- Pattern Guard P/R/F1.}
Which fault categories does the static screen actually catch on the E2
mutant population?

\paragraph{E8 --- Generalisation.}
Do SpecIR adapters carry the loop beyond SOFL?  Conf for M, B1, B2 on
SOFL/FSF, Mini-Z, and Mini-StateMachine (adapter $\to$ SpecIR $\to$ shared
pipeline).
E8b expands the Mini-Z and Mini-StateMachine corpora to $n{=}30$ tasks each
(\texttt{run\_e8b\_expanded\_v1}); E8c reports full B1/B2/M on 40
real-derived HumanEval/MBPP tasks.

\paragraph{E9 --- Failure Analysis.}
What remains after $K=3$?  Taxonomy of non-accepted tasks under mode~M.

\paragraph{E10 --- Random Benchmark.}
Does M's advantage persist without overlap filtering?  100 uniformly sampled
tasks $\times$ modes B1, B2, M (Section~\ref{sec:exp:e10}).

\paragraph{E12 --- Multi-Seed Stability.}
Does the M--B2 ranking flip under LLM stochasticity?  30 stratified tasks
(from \texttt{benchmarks/e12\_stratified\_30.json}) $\times$ seeds
$\{0,1,2\}$ $\times$ modes B2, M (\texttt{run\_e12\_stability\_v1}).
Per-mode mean, standard deviation, win rate, and Friedman test across seeds
are reported in Section~\ref{sec:results:e12}.

%=============================================================================
\section{Statistical Protocol}
\label{sec:exp:stats}
%=============================================================================

All pairwise comparisons involving per-task binary outcomes (strict success) use the
\emph{Wilcoxon signed-rank test}; latency comparisons use the \emph{Mann-Whitney U test}.
The full statistical protocol comprises:

\begin{itemize}
  \item Paired Wilcoxon signed-rank test for conformance comparisons.
  \item Mann-Whitney U test for latency comparisons.
  \item Holm-Bonferroni correction applied across all pairwise mode comparisons.
  \item Cliff's $\delta$ as non-parametric effect size.
  \item Bootstrap 95\% confidence intervals ($n=1000$ resamples).
  \item Significance threshold $\alpha = 0.05$.
\end{itemize}

Effect sizes are interpreted as: $|\delta| < 0.147$ negligible,
$0.147 \le |\delta| < 0.33$ small, $0.33 \le |\delta| < 0.474$ medium,
$|\delta| \ge 0.474$ large.
We apply Holm-Bonferroni correction for E1 all-mode comparisons.

\paragraph{Post-hoc power analysis (REV-5).}
With $n{=}120$ paired tasks, observed Cliff's $\delta$ for M vs.\ B1 conformance
is $+0.036$ (Wilcoxon $p{=}0.010$, Holm $p{=}0.075$): the benchmark is
marginally powered for that comparison.
For M vs.\ B2, $\delta{\approx}0.0002$ (Holm $p{=}0.811$): the observed gap is
negligible and post-hoc power is ${\approx}2.5\%$ at $\alpha{=}0.05$---far below
the conventional 80\% target.
Detecting an effect of this size would require $n{\gtrsim}10^{5}$ paired tasks
for 80\% power at $\alpha{=}0.05$ (\texttt{power\_analysis.json}); an
$n{=}240$ extension was therefore not pursued.
We report per-task win rate (6/120), overlap stratification (E3, E10),
multi-seed stability (E12), and VerifierLoop-FSF (B6) rather than treating the
aggregate M--B2 conformance delta as a headline claim.
Analysis script: \texttt{power\_analysis.py} $\to$
\texttt{data/processed/power\_analysis.json}.

\subsection{Multi-Seed Protocol (E12)}
\label{sec:exp:e12}

E12 selects 30 tasks stratified by overlap tertile and guard-complexity label
(\texttt{scripts/select\_stratified\_subset.py}) and re-runs modes B2 and M with
three independent LLM seeds while holding $T_\text{gen}$, $T_\text{repair}$, and
$K$ fixed.
The primary stability outcomes are: (i)~whether M's mean conformance exceeds B2's
in every seed; (ii)~per-task win rate M vs.\ B2 pooled across seeds; and
(iii)~Friedman test across seeds for ranking reversals.
This supplements the single-repeat E1 matrix without re-running all 120 tasks
$\times$ 10 modes at three seeds.

%=============================================================================
\section{Threats to Validity (Preview)}
\label{sec:exp:threats-preview}
%=============================================================================

Threats follow the four-category framework of empirical software
engineering~\citep{ammann2016testing,woodcock2009formalsurvey}; full discussion appears
in Chapter~\ref{ch:discussion}.

\paragraph{Construct validity.}
Formal conformance is approximated via Z3-generated concrete cases, not full theorem
proving; Theorem~\ref{thm:far-bound} quantifies the FAR bound under finite
witness budgets.
The 14-pattern guard catalogue (RF01--RF14) may miss novel fault classes; E7 reports
per-pattern P/R/F1 to expose gaps.
PDR uses first-order mutation operators; higher-order and real-world faults may differ.

\paragraph{Internal validity.}
LLM non-determinism is mitigated by $T_\text{repair}=0.0$, fixed seeds, bootstrap CIs over
tasks, and E12 multi-seed stability on a 30-task stratified subset.
The extended B3--B5 baselines use dedicated \texttt{config\_for\_mode} branches
validated before merge.
Reference implementations are auto-generated via \code{reference\_gen.py} with
hand-crafted overrides for core tasks.
Benchmark construction by the same authors introduces possible tuning bias, mitigated
by ablation (A1--A3), sensitivity analysis, E10 (no overlap filter), and the
independent E8 subset.
The Z3 overlap filter selects tasks where SMT-based approaches are advantaged.

\paragraph{External validity.}
Primary evaluation uses 120 synthetic numeric-domain tasks; E8 validates generalisation
on Mini-Z, Mini-StateMachine, and real-derived tasks at smaller scale.
Implementations are Python; results may not transfer to other languages or LLMs.

\paragraph{Conclusion validity.}
Holm-Bonferroni correction, Cliff's~$\delta$, and bootstrap 95\% CIs are applied
throughout.
The strict-success null result (M vs.\ B1/B2, uncorrected $p \approx 0.92$;
Holm $p = 1.000$) reflects the conjunctive acceptance gate design, not method
equivalence on conformance.
The full prevention evaluation supersedes preliminary spec-confusion results
($n=40$); the non-monotonic B2/M ordering reflects complementary detection
mechanisms.

With the protocol fixed, the next chapter reports what those experiments
actually show---mechanism first, then overall effectiveness.

% -----------------------------------------------------------------------------
\section*{Chapter Summary}
\addcontentsline{toc}{section}{Chapter Summary}
% -----------------------------------------------------------------------------

Nine core experiments (E1--E9) plus E10 (random benchmark) and E12
(multi-seed stability) test the three Tier-1 claims across five research
questions on a 120-task hard benchmark with ten compared modes.
Prevention is measured by PDR and FAR alongside formal conformance and latency;
statistical comparisons use Holm-Bonferroni-corrected non-parametric tests.
Threats are previewed here and fully analysed in Chapter~\ref{ch:discussion}.
