%% ch08_discussion.tex  —  Chapter 8: Discussion and Threats to Validity
%% HSP-Agile CCF-B Report, East China Normal University, 2026
%% Step 1 polish staging: _pipeline/step1_polish/

\chapter{Discussion and Threats to Validity}
\label{ch:discussion}

Chapter~\ref{ch:results} reports the measurements.
This chapter interprets them for sprint practice: when the conjunctive release
bar justifies the full loop, where claims stop, and which validity threats
remain in view.
A short preview appeared in Section~\ref{sec:exp:threats-preview}; the
analysis below is the authoritative reading.


%=============================================================================
\section{Interpretation of Core Findings}
\label{sec:disc:interpretation}
%=============================================================================

Acceptance safety is the lead finding.
Conjunctive release
\[
  \mathrm{Accept}(c,t) \iff \mathrm{Conf}(c,t)=1 \land \mathrm{Screen}(c)=\mathrm{pass}
\]
binds witness conformance to a structural screen
(Equation~\eqref{eq:accept}).
A candidate that greens a busy unit suite can still fail Accept when a
critical pattern remains, or when a thin residual region was never exercised.

Two controlled results support that reading.
Under feedback isolation (E6), scenario-indexed Semantic Feedback IR lifts
mean Conf.\ on headroom tasks: from 79.1\% (\texttt{test\_only}) to 86.9\%
(\texttt{semantic\_ir}) on the full panel (${+}7.7$\,pp supporting
aggregate), and ${+}8.2$\,pp on the broad headroom set where test-only
Conf.${<}1$ ($n{=}114$; W/L/T $14/4/96$; CI $[2.3,14.4]$).
The collapsed subset with test-only Conf.${=}0$ ($n{=}13$) explains the
rescue regime (${+}90.7$\,pp; 13/13 wins) but is too small to headline alone.
The partial-failure band ($0{<}$test-only Conf.${<}1$, $n{=}101$) does not
favour typed IR (${-}2.5$\,pp; CI $[-5.8,0.2]$), because failures there are
often multi-scenario or arithmetic partial mismatches in which a single
scenario label can mis-steer repair.
All fourteen Conf.\ wins are catch-all/\texttt{others} rescues; none are
ordering-only uniqueness wins.
Among the 18 decisive (non-tie) tasks, Semantic IR wins 14/18 (77.8\%;
Wilson 95\% CI $[54.8,\,91.0]$).
The partial-failure band ($0{<}$test-only Conf.${<}1$, $n{=}101$) does not
favour typed IR (${-}2.5$\,pp; CI $[-5.8,0.2]$); keep B2 rather than forcing
Semantic IR there.
That profile matches the primary vignette: Reject and Pass tests stay green
while the residual \texttt{OTHERWISE} branch still returns Pass instead of
Review.
Under implementation screening (E2), the industrial slice ($n{=}132$) moves
FAR from 56.8\% (B2/A2) to 32.6\% (M)---the primary prevention claim
(bootstrap 95\% CI on the FAR drop: $[-31.8,\,-17.4]$\,pp; excludes~0).
Pooled over $n{=}852$ that lift appears as FAR 8.8\% to 5.0\% (PDR 91.2\% to
95.0\%), but HardSynthetic mutants do not exercise Screen and pooled FAR 5.0\%
is \emph{not} the release claim.

Equal-$K$ Conf.\ ranking places M\_eq at 100.0\% mean Conf.\ against B2's 97.5\%
(${+}2.5$\,pp; $3/0/117$) when both modes share $K{=}3$.
The three wins are a matched-budget Conf.\ check, not a substitute for E6's
within-M feedback isolation.
Near-ceiling Conf.\ rankings on the strengthened $K{=}5$ bundle (fixed-oracle
E1/E10) and the multi-seed E12 matrix are saturation context rather than a
Conf.\ crown: B2 reaches 100\% Conf.\ on every E12 seed while M averages
99.7\%.
Aggregate E1/E10 place M at or above B2; seed stability slightly favours B2.
The practical reading is conditional: prefer B2 for mean Conf.\ and
latency; escalate to~M when $\mathrm{Accept}{=}1$ or a contractual FAR budget
exists \emph{and} repair headroom remains
(Table~\ref{tab:deployment-boundary}).
Do not read the industrial Screen lift (FAR 56.8\% to 32.6\%) as attainment of
FAR${\le}5\%$ on that mutant set.
When B2 already achieves Accept${=}1$ (or 100\% mean Conf.\ on the release
suite), do not escalate---the fuller loop can regress under strong models
(GPT-4o/Claude industrial-pattern: B2 100.0\% vs.\ M 95.5\%).
The controlled causal estimate for SIFR remains E6 on BENCH-120; Gemini
hard-seed and public-corpus runs bound saturation and transfer, not a second
${+}7.7$\,pp headline.

\paragraph{Failure modes and escalation limits.}
Four negative results bound when \emph{not} to escalate: partial-failure band
(${-}2.5$\,pp; do not force SIFR); GPT-4o/Claude industrial-pattern regression
(B2 100.0\% vs.\ M 95.5\%); E14 length-matched traces beating bare Semantic IR
(85.4\% vs.\ 75.1\%); and HTE precision 0.167 (checklist only, not a
predictor).
Table~\ref{tab:deployment-boundary} records these limits beside the positive
triggers in Section~\ref{sec:results:rq4}.


\paragraph{When baselines already saturate.}
Industrial-pattern, desensitized, and several harvest slices often tie B2 with
M\_eq at 100\% mean Conf.
When test feedback already clears every witness, B2 is the correct deployment
choice.
Value concentrates on contracts with logical traps---catch-all fail-closed
residuals, MFA step-up bands, ordered overlaps---where a green dashboard can
still hide an escaped defect (Section~\ref{sec:results:acl-case}).


\paragraph{Strengthened mode~M and measurement.}
Default mode~M on the fixed-oracle ranking runs combines several engineering
axes distinct from the single-factor E6 claim: best-effort
$\arg\max_k\mathrm{Conf}$ selection, length-matched
\texttt{execution\_trace\_matched} repair feedback (E14), advisory pattern
gate with hard block reserved for critical severity (E17), $K{=}5$, and
\texttt{formal\_max\_cases}${=}32$.
E6 continues to isolate \texttt{semantic\_ir} versus unstructured
\texttt{test\_only} under controlled $K{=}3$ comparison.
A prior others-witness bug (catch-all region encoded as True without
$\neg$prior guards) inflated false failures on historical E1; fixing
first-match others witnesses is a measurement correction that enables fair
evaluation and is disclosed under threats
(Section~\ref{sec:disc:threats}).


%=============================================================================
\section{The Quality--Pass-Rate Paradox}
\label{sec:disc:paradox}
%=============================================================================

Mean formal conformance and strict success answer different questions.
$\mathrm{Conf}$ is continuous over a finite witness set: a candidate that
satisfies nine of ten cases contributes $0.9$ to the average.
Strict success is binary and conjunctive: $\mathrm{Accept}$ demands full
witness conformance \emph{and} a clear critical pattern screen.
On fixed-oracle BENCH-120 the two track closely for B1/B2/M, with only a
narrow M--B2 Conf.\ margin.

When release quality is measured by escaped defects rather than average
agreement, prevention metrics dominate the reading.
The conjunctive gate can still reject candidates that pass all witnesses yet
exhibit high-severity pattern violations---candidates that B1/B2's
single-criterion gate would accept.
On industrial-style mutants, Screen cuts FAR from 56.8\% to 32.6\%---the
primary prevention claim.
Pooled over the implementation-screening set that yields FAR 8.8\% to 5.0\%,
but HardSynthetic mutants do not exercise Screen and pooled FAR 5.0\% is
\emph{not} the release claim.
Earlier snapshots in which Conf.\ and Strict diverged under the buggy
others-witness oracle are superseded; the enduring lesson is conjunctive
selectivity plus honest measurement, not a single Strict headline.


%=============================================================================
\section{Implications for Practice}
\label{sec:disc:implications}
%=============================================================================

\subsection{Sprint Integration and Deployment Heuristic}

\textsc{SgDP} sits between test-feedback (fast, spec-unaware) and full
deductive verification (slow, annotation-heavy): a heuristic for FSF-style
ordered-guard workflows pending live industrial replication.
On fixed-oracle E1, mode~M and B2 incur comparable mean wall-clock latency
($10{,}457$\,ms vs.\ $9{,}576$\,ms) at similar mean llm\_calls
(${\approx}1.16$ vs.\ ${\approx}1.18$).
When mean Conf.\ and latency suffice, Table~\ref{tab:deployment-boundary}
prescribes B2; escalate to~M only when $\mathrm{Accept}{=}1$ or a contractual
FAR budget exists \emph{and} repair headroom remains.
Screen helps on industrial-pattern mutants but does not attain FAR${\le}5\%$
on that slice.
The conjunctive predicate additionally yields an auditable decision
record---which witnesses and patterns were decisive---rather than a bare
pass/fail bit~\citep{harman2004maff,beck2004xp}.

\subsection{The Formal Checker as a Lightweight Oracle}

The Z3-based conformance checker occupies a middle ground between no formal
verification and full deductive tools such as Dafny~\citep{leino2010dafny}
or Coq~\citep{woodcock2009formalsurvey}.
It provides \emph{bounded} guarantees over a finite, specification-derived
witness suite at less than $100$\,ms per task---negligible relative to
pipeline latency---without requiring verification annotations or dedicated
verification engineers.
Ablations under the fixed oracle show that removing the checker (A1) or the
repair loop (A3) costs substantial Conf.; removing the advisory pattern gate
(A2) costs none on Conf.\ at ceiling, so the screen's deployment rationale
remains prevention (E2), not Conf.\ ranking.

\subsection{Static Screen Selectivity and Replaceability}

For most smells, advisory warnings are preferable to hard rejections,
following severity-modulated static-analysis practice
\cite{hovemeyer2004findbugs,ayewah2007findbugs}, with hard rejection reserved
for RF07-class shortcuts; E17 evaluates that advisory operating point.
The screen is a configurable slot
(Section~\ref{subsec:method:pluggable-screen}): new failure modes observed in
production can be added as smells without retraining the synthesis LLM, or the
entire PoC catalogue can be swapped for a CI analyser or LLM review agent that
emits the same severity-tagged vetoes.
E7's per-smell F1 guides catalogue expansion; it does not freeze fourteen
rules as the only admissible $\mathrm{Screen}$.
The fourteen-rule catalogue is an evaluated implementation choice,
not evidence of completeness.  E2 includes semantic mutants beyond direct
rule matches, and the A2 ablation separates the screen's prevention role from
the witness/repair contribution to conformance.

\paragraph{Finite-sample FAR envelope versus operational FAR.}
Proposition~\ref{thm:far-bound} is a finite-sample envelope under
independent witness draws; the deployed $\Phi_i$ suite is deterministic and
does not assume that independence.
Release decisions should cite empirical E2 FAR (industrial slice 56.8\% to
32.6\%; pooled 5.0\% vs.\ B2 8.8\%), not the symbolic
$(1-\varepsilon)^n$ expression.


%=============================================================================
\section{Generalisation Boundaries}
\label{sec:disc:generalisation}
%=============================================================================

The framework assumes deterministic, enumerable guards encodable as SMT
constraints---FSF-style ordered scenarios, finite-state transition guards,
and simplified Z/TLA+ preconditions over bounded domains.
Generalisation experiments (E8) show that M outperforms B1 across SOFL/FSF,
Mini-StateMachine, and Mini-Z with adapter-level changes, supporting a
framework claim rather than a single-notation artefact.
Absolute Conf.\ on expanded Mini adapters remains low because foreign syntax
constrains generation; real-derived HumanEval/MBPP transfers are stronger,
with B2 leading on the sparser HumanEval-FSF subset.

\paragraph{When test-feedback is enough.}
B2 matches or beats the full pipeline when guards barely overlap (sparse
HumanEval-FSF and E11 external slices), branching is shallow
(typically ${\le}4$ branches), or each branch returns a simple scalar so that
ordinary failures already expose assignment mistakes.
Under those conditions the lightweight loop is the rational default.
M's contribution on the hard benchmark is conditional: E6 typed
feedback under headroom, E2 prevention on industrial-style mutants, and
conjunctive $\mathrm{Accept}$ when those release bars apply.

\subsection{Deployment Boundary Conditions}
\label{sec:disc:deployment}

Table~\ref{tab:deployment-boundary} summarises the decision: match feedback
depth to the release bar, not to overlap tertiles alone (canonical E3 is
non-monotone).
The pre-hoc HTE diagnostic remains deliberately conservative: the best
inspected tree over hard-task metadata reaches precision 0.167 and recall
0.786 against a 0.117 base positive rate, while external summaries are mostly
saturated, tied, or aggregate-only.
We treat HTE as a checklist policy rather than as a validated predictor.

\begin{table}[t]
  \centering
  \caption{\textbf{Deployment boundary:} release bar + headroom (not overlap).
    Prefer B2 for mean Conf./latency; escalate to~M when Accept or a contractual
    FAR budget applies \emph{and} repair headroom remains.
    Within M, prefer \texttt{semantic\_ir} for localisation/audit and
    \texttt{execution\_trace\_matched} for mean Conf.\ (E14).
    Overlap tertiles (E3) are non-monotone---not a deployment trigger.}
  \label{tab:deployment-boundary}
  \footnotesize
  \renewcommand{\arraystretch}{1.12}
  \begin{tabular}{@{}p{0.28\linewidth}p{0.32\linewidth}p{0.32\linewidth}@{}}
    \toprule
    \textbf{Signal} & \textbf{Prefer B2} & \textbf{Prefer M} \\
    \midrule
    Default & Mean Conf.\ / latency & --- \\
    Release bar & Mean Conf.\ sufficient & $\mathrm{Accept}{=}1$ \\
    Prevention & No FAR contract & Contractual FAR budget \\
    Headroom & Saturated endpoint / ties & Collapsed test-only Conf.\ (E6) \\
    Partial Conf. & $0{<}$test-only Conf.${<}1$ & Do not force SIFR \\
    Strong-model Accept & B2 already Accept${=}1$ & Do not escalate \\
    Feedback render & Unstructured OK if Conf.\ holds & IR (audit) / trace (Conf.) \\
    Seed stability & B2 first every E12 seed & E1/E10 M$\ge$B2 aggregate \\
    Hybrid cost & M\_lite not a shortcut (E18) & Full IR$+$gate integration \\
    Overlap tertile & Non-monotone (E3/E10) & \textbf{Not} alone \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Hybrid baselines.}
For mean Conf.\ alone, prefer B2 over verifier-loop hybrids (E13).
M\_lite (E18) and B6 do not reproduce M's prevention profile: as formal-only
modes without Screen, B6 prevention equals B2/A2 by design (HardSynthetic FAR
0\%; industrial B6 unavailable in \texttt{prevention\_full\_v1}), and we do not
report B6 FAR without \texttt{prevention\_full\_v1} rows.
Deploy full IR$+$gate integration or stay on B2.

The framework does not yet encode temporal LTL/CTL properties, infinite-domain
arithmetic beyond LIA, or probabilistic semantics.
Industrial replication on practitioner SOFL/FSF projects remains necessary
before quantitative production claims~\citep{li2022qrs_companion,pohl2010requirements}.


%=============================================================================
\section{Limitations}
\label{sec:disc:limitations}
%=============================================================================

\paragraph{Benchmark scale and construction.}
BENCH-120 balances overlap-filtered discriminative tasks against LLM API cost.
Post-hoc power for the tiny fixed-oracle M--B2 Conf.\ gap is low
($|\delta|{\approx}0.014$); doubling $n$ would not materially change
inference.
We emphasise per-task win rate, equal-$K$ Conf.\ ranking, E10
(unfiltered), E12 (multi-seed), and prevention metrics alongside aggregate
means.
Author-construction bias is mitigated by the real-priority micro-benchmark
($n{=}30$), GitHub harvest ($n{=}48$; B1 headroom then B2${=}$\texttt{M\_eq}),
HKCA09, and desensitized published-industrial pilots---still not live vendor
traces (NDA intake pending; public harvest $\neq$ proprietary dump).
External impl-screening shows GitHub Screen transfer
(FAR 60.4\%$\rightarrow$11.5\%) beside published-industrial saturation
(0 Screen-hits); industrial-132 remains the primary Screen claim.
Industrial specifications may add state, external services, or composite
scenario structure absent from the template family.
The E2 prevention estimate is conditional on an author-selected first-order
mutant distribution derived from SOFL fault patterns.  Although the operators
extend beyond direct RF01--RF14 matches, the study does not establish screen
completeness, clean-repository false-rejection rates, or repository-scale
safety.  The 5.0\% pooled FAR is an empirical result for this constructed
population, not a universal release guarantee.
Finite witnesses approximate Conf.; Conf.${=}1$ is not a completeness claim
over the full input space.
Accept is sound only relative to a faithful SpecIR$\to$SMT encoding of the
guards under evaluation; encoding bugs (such as the disclosed others-witness
fault) change measured Conf.\ and must be treated as measurement corrections.
The Screen catalogue is a configurable instance, not a completeness proof over
defect classes.

\paragraph{Overlap filter and claim scope.}
The Z3 overlap filter retains tasks where first-match ordering defects are
possible; without overlap, precedence bugs are rare and test-feedback is
advantaged.
E10 removes the filter; E3 stratifies by overlap tertile.
Claims are scoped to overlap-intensive ordered-guard specifications, not
universal code generation.

\paragraph{Model endpoints.}
Primary results use \texttt{ecnu-plus} (Qwen-family); E16/E6-max use
\texttt{ecnu-max} (DeepSeek-family); commercial replication uses n1n
\texttt{gpt-4o}, \texttt{claude-sonnet-4-6}, and \texttt{deepseek-v3.2}
(Section~\ref{sec:results:n1n}).
Cross-endpoint evidence shows that headroom, rather than model prestige alone,
determines the value of added repair structure.
On industrial-pattern tasks, B2 reaches 100\% Conf.\ under GPT-4o and Claude
while M reaches 95.5\%; Gemini hard-seed repair retains a ${+}26.5$\,pp
structured-over-test-only contrast (CI excludes zero).
Regression guard: when B2 already achieves Accept${=}1$, do not escalate to~M.
The controlled headroom evidence (+8.2\,pp on test-only Conf.${<}1$; all-task
+7.7\,pp as supporting aggregate) remains primary-endpoint evidence on
BENCH-120; the commercial and Gemini runs establish saturation and transfer
boundaries, not a second causal headline.

\paragraph{Witness coverage and SMT theory class.}
The checker generates a small number of witnesses per branch inside a
parameterized integer box (default $[-5,20]$; overridable per task or CLI).
Table~\ref{tab:smt-scalability} shows that widening to $[-100,100]$ or
$[-1000,1000]$---and probing Z3 Reals---does not create a latency cliff on
the real-priority micro-benchmark's linear guards.
Denser boundary perturbation would raise fault-detection sensitivity at higher
latency; the current regime likely underestimates M's advantage versus a denser
suite.
The honest limit is theory class: nonlinear arithmetic, large bit-vectors, or
unbounded heap shapes can make SMT return \texttt{unknown} or time out.

\subsection{Hybrid Fallback: SMT Timeout $\to$ Fuzzing + Heuristics}
\label{sec:disc:smt-hybrid}

When the witness generator hits a wall, \textsc{SgDP} should not silently skip
the region.
The prescribed contingency---to be instrumented as a first-class pipeline mode
in follow-on work---is:

\begin{enumerate}
  \item \textbf{Budgeted SMT first.} Attempt $\Phi_i$ with the configured
        domain box and a hard wall-clock budget.
  \item \textbf{Fuzz the residual.} If SMT fails, sample toward guard
        boundaries (perturb neighbouring witnesses; flip relational operators;
        draw from empirical traffic if available).
  \item \textbf{Heuristic region seeds.} For ordered guards such as
        \code{compute\_tax}, inject geometric overlap corners from
        Figure~\ref{fig:defect-vs-test} even when the solver cannot prove them.
  \item \textbf{Degrade Accept, do not fake it.} Fuzz witnesses are labelled
        \texttt{heuristic} in the Semantic Feedback IR; conjunctive
        $\mathrm{Accept}$ may stay advisory until SMT recovers.
\end{enumerate}

\noindent
Figure~\ref{fig:smt-domain-latency} motivates treating the hybrid as a
contingency rather than the default: on the FSF LIA fragment that dominates
our corpora, pure SMT stays in tens of milliseconds after domain widening.

\paragraph{Repair budget.}
$K$ is fixed across tasks.
Harder tasks may benefit from larger $K$; simpler tasks waste iterations.
Adaptive allocation based on observed Conf.\ improvement per round remains
future work.

\paragraph{Historical residual taxonomy.}
E9 failure patterns were collected under the pre-fix others-witness oracle
and are historical archive only (Appendix~\ref{app:stress},
Section~\ref{sec:app:e9}); they are not current Strict evidence under the
fixed oracle.


%=============================================================================
\section{Comparison to Related Approaches}
\label{sec:disc:related}
%=============================================================================

\paragraph{DafnyBench and full deductive verification.}
DafnyBench~\citep{loughridge2024dafnybench} evaluates LLM synthesis of Dafny
programs with machine-checked proofs.
Full deductive verification provides stronger guarantees than bounded Z3
checking but requires annotations absent from FSF.
HSP-Agile applies directly to SOFL/FSF inputs without verification
annotations, occupying a complementary position on the verification-cost
spectrum.

\paragraph{CEGIS-inspired approaches (Sun et al., Clover).}
Clover~\citep{sun2024clover} uses proof-obligation counterexamples to guide
LLM repair in Dafny.
HSP-Agile likewise turns witnessed failures into repair prompts, but the
counterexamples come from Z3-driven tests derived from FSF scenario semantics.
The checker is lighter-weight and the guarantees correspondingly weaker.

\paragraph{VerifierLoop-style FSF baseline (B6).}
Mode~B6 implements SMT witness checking inside the repair loop with structured
counterexample prompts, equal $K$ and witness budget as M, but without
Semantic Feedback IR or pattern guard.
On stratified and full hard exports, aggregate Conf.\ rankings are
seed-sensitive near ceiling (cf.\ E12); the irreducible increment of full~M
over such hybrids remains conjunctive $\mathrm{Accept}$ together with
PDR/FAR on industrial-style mutants---and the E6 typed-IR gain when operating
inside M's repair loop under headroom.
B6 prevention equals B2/A2 by design when Screen is off (HardSynthetic FAR
0\%; industrial B6 unavailable in \texttt{prevention\_full\_v1}); cite
BENCH-120 Conf.\ (B6 81.1\% vs.\ M\_lite 79.0\%) for conformance only, not
B6 FAR without \texttt{prevention\_full\_v1} rows.
Clover remains the verified CEGIS reference for Dafny; B6 is our FSF
adaptation, not a drop-in of a third-party FSF product.

\paragraph{Standard LLM baselines (B1 and B2).}
Adoption should follow the deployment table and per-task evidence rather than
a blanket ranking across industrial corpora~\citep{dakhel2023copilot,chen2021codex}.
HSP-Agile targets quality gates where downstream formal costs dominate, not
real-time code completion.


%=============================================================================
\section{Threats to Validity}
\label{sec:disc:threats}
%=============================================================================

We follow the standard split into construct, internal, external, and
conclusion validity
\cite{ammann2016testing,wohlin2012experimentation}.

\paragraph{Internal validity --- author-built screens and corpora.}
The Stage~4 PoC smell catalogue (RF01--RF14) was written by the same team that
designed the pipeline, creating a risk of designer bias toward faults expected
on BENCH-120 and SOFL-derived mutants.
Ablations (A2), advisory mode (E17), and E7 per-smell F1 make the PoC
measurable but do not prove the fourteen rules unique or unbiased.
Mitigations: treat RF01--RF14 as a pluggable instance of $\mathrm{Screen}$
(Section~\ref{subsec:method:pluggable-screen}); fix seeds and
$T_\text{repair}{=}0$ for repair calls; report E12 multi-seed stability;
include baselines B3--B6 and equal-$K$ Conf.\ ranking.
Residual risk remains: a different smell pack or review agent could shift
which candidates Accept rejects.

\paragraph{External validity --- first-match is not every business rule.}
Primary tables lean on synthetic overlap-heavy FSF tasks plus curated
micro-benchmarks.
Enterprise logic is often stateful, temporal, or table-driven in ways that do
not reduce cleanly to ordered first-match guards.
Adapters (E8) and the real-priority micro-benchmark
(Section~\ref{sec:results:real-priority}) show transfer when the contract
\emph{is} FSF-shaped; they do not show that arbitrary production rules can be
rewritten as SpecIR without semantic loss.
Claims are scoped to ordered-guard / FSF-style workflows; live vendor
replication remains future work.
Headroom-filtered external hard seeds
(\texttt{run\_ext\_hardseed\_e6\_v2}: \texttt{wrong\_relop} $n{=}79$,
${+}20.6$\,pp; HKCA09 gemini $n{=}17$, ${+}18.8$\,pp) support protocol transfer
under test-only Conf.${<}1$; they do not solve external validity on their own.

\paragraph{Construct validity --- Conf., PDR, and FAR.}
$\mathrm{Conf}$ approximates oracle agreement on a finite witness set
$\mathcal{D}(t)$; $\mathrm{Conf}{=}1$ is not a completeness claim over the full
input space.
$\mathrm{PDR}$/$\mathrm{FAR}$ score rejection of first-order mutants under the
conjunctive Accept gate.
Neither construct covers higher-order faults, environment faults outside the
FSF signature, timing/concurrency, or inputs outside the SMT domain box.
Accept is sound only relative to a faithful SpecIR$\to$SMT encoding of the
guards under evaluation.
Theorem~\ref{thm:far-bound} does not license operational release decisions;
cite empirical E2 FAR instead (industrial slice first, then pooled aggregate).
\emph{Disclosure:} a prior others-witness bug inflated false failures on
canonical E1; primary B1/B2/M aggregates cite the fixed-oracle
\texttt{run\_e1\_m\_win\_v2} only
(Section~\ref{sec:disc:interpretation}).

\paragraph{Conclusion validity --- seeds, Strict, and screen ablations.}
Authoritative E1 cites a single generative seed (\texttt{run\_e1\_m\_win\_v2},
repeat~0) with win rate and means; fixed-oracle E12 re-ran all 120 tasks
across three seeds for B1, B2, and M.
B2 leads every seed at 100.0\% mean Conf.; M averages 99.7\% (one failure on
seed~1).
Both are near ceiling; we report win rate and means for fixed-oracle E1
(M vs.\ B2: $2/120$ wins, ${+}1.7$\,pp) without a Holm claim on that narrow
gap.
Full E2 supersedes preliminary $n{=}40$ prevention numbers.
Under the fixed oracle, Strict tracks Conf.; the historical uniform
${\approx}5\%$ Strict on pre-fix E1 was largely a measurement artefact.
A2's Conf.\ delta under the advisory gate at ceiling is zero on the fixed
ablation; the screen's primary evaluation remains E2 PDR/FAR, with E17 as the
latency-friendly advisory alternative when a hard gate is too strict.


%=============================================================================
\section{SpecIR Scope Assumptions and Future Extensions}
\label{sec:disc:specir-scope}
%=============================================================================

The current SpecIR design (Definitions~\ref{def:specir}--\ref{def:specir-wf})
makes explicit assumptions that bound framework generality.
Table~\ref{tab:specir-scope} maps each assumption to a natural extension;
Section~\ref{sec:conc:future} prioritises these directions.

\begin{table}[t]
  \centering
  \caption{\textbf{SpecIR scope assumptions and extension paths.}
    Each row states a current design choice, its limitation, and a research
    direction beyond the HSP-Agile evaluation.}
  \label{tab:specir-scope}
  \footnotesize
  \renewcommand{\arraystretch}{1.15}
  \begin{tabular}{@{}p{0.22\linewidth}p{0.30\linewidth}p{0.38\linewidth}@{}}
    \toprule
    \textbf{Assumption} & \textbf{Limitation} & \textbf{Extension} \\
    \midrule
    Deterministic guards &
    Cannot express probabilistic thresholds or noisy sensors &
    Probabilistic guards with confidence intervals \\
    First-match ordering &
    No temporal ``always/eventually'' constraints &
    LTL/CTL adapters over transition systems \\
    Single oracle $g_t$ &
    No distributed or multi-service contracts &
    Compositional oracles with interface contracts \\
    Linear integer arithmetic (LIA) &
    Non-linear and floating-point guards fail encoding &
    NIA/QF\_FP backends with staged fallback \\
    \bottomrule
  \end{tabular}
\end{table}

These assumptions are deliberate scoping choices for a first framework
instantiation.
E8 shows that adapter-level extensions preserve the shared pipeline; temporal
and probabilistic semantics would need new adapter and witness modules while
retaining the Semantic Feedback IR and acceptance predicate.

\input{chapters/ch08_industrial_case}
