%% Industrial decision vignette — included from ch08_discussion.
%% Step 1 polish staging: _pipeline/step1_polish/ch08_industrial_case.tex

\section{Industrial Decision Vignette}
\label{sec:disc:industrial-case}

Ordered-guard contracts appear outside synthetic benchmarks.
A telecom-style alert table in \texttt{benchmarks/industrial\_sofl.json}
encodes overlapping severity guards in which ``link down'' must preempt
``high latency'' when both predicates hold---the same first-match
\emph{ordering} discipline as the secondary overlap vignettes of
Chapter~\ref{ch:intro} (\code{classify\_signal} / tax), now with industrial
wording.
Catch-all residual mismatch remains the primary Conf.\ mechanism on
BENCH-120 (E6 headroom strata); this vignette stresses preemption witnesses
and deployment choice under saturation.

On the $n{=}31$ industrial-pattern corpus, commercial endpoints already
saturate mean Conf.\ under test-feedback: B2 reaches 100\% under gpt-4o and
Claude while M sits at 95.5\% (Table~\ref{tab:cross-model-industrial}).
The Wave-2 desensitized published-industrial pilot ($n{=}28$ reconstructions
from public railway crossing, interlocking, Agile-SOFL ATM, and book/IET
cases; \texttt{run\_desens\_real\_sofl\_v1}; Table~\ref{tab:desens-real-sofl})
extends the reading under \texttt{ecnu-plus}: B1, B2, and equal-$K$ M\_eq all
reach 100\% Conf.

The practical rule follows Table~\ref{tab:deployment-boundary}.
When mean Conf.\ before merge is the release bar, B2 is the correct default.
Escalate to~M only when the sprint requires conjunctive
$\mathrm{Accept}{=}1$ (every SpecIR witness and a clear critical pattern
screen) or a prevention target FAR${\le}5\%$ on mutant screening (E2,
industrial slice first).
Overlap density is not the trigger: E3 tertiles on BENCH-120 are
non-monotone, and both industrial slices already favour B2 on Conf.

\paragraph{What saturation leaves open.}
Table~\ref{tab:defect-evidence-map} clarifies the complementary roles.
E6's typed-IR gain is a scenario-indexed rescue under repair headroom on
BENCH-120; ordering and preemption defects still motivate $\Phi_i$ witnesses
and E2 mutants on the industrial slice.
Saturated industrial Conf.\ supports the B2 default for mean quality;
Accept and PDR/FAR remain the reasons to escalate when release bars demand
them.

These corpora are curated industrial-\emph{pattern} SOFL and desensitized
published reconstructions, not live vendor traces.
They support the deployment heuristic; quantitative vendor KPIs remain
future work pending an instrumented Agile-SOFL pilot.
