%% ch09_conclusion.tex  —  Chapter 9: Conclusion and Future Work
%% HSP-Agile CCF-B Report, East China Normal University, 2026
%% Ye Xujun
%% Step 1 polish staging: _pipeline/step1_polish/

\chapter{Conclusion and Future Work}
\label{ch:conclusion}

Acceptance safety is a release decision: ship only when every
specification-derived witness matches \emph{and} a structural screen clears
($\mathrm{Accept}\Leftrightarrow\mathrm{Conf}{=}1\wedge\mathrm{Screen}{=}
\mathit{pass}$).
On ordered-guard contracts that decision matters because a busy Reject/Pass
suite can stay green while a catch-all residual still returns the wrong
tier---the classifier vignette in which \texttt{OTHERWISE} must yield
\texttt{Review} yet an LLM candidate emits \texttt{Pass}.
HSP-Agile offers headroom-conditioned catch-all repair under a SpecIR
first-match oracle, plus a conditional dual-gate acceptance check for
ordered-guard specifications:
$\mathrm{Accept}\Leftrightarrow\mathrm{Conf}{=}1\wedge\mathrm{Screen}{=}\mathit{pass}$.
The contribution is an operational acceptance-checking stack---not a new SMT
solver or a proof system---realised on SOFL/FSF with a pluggable screen.
As a formal-engineering method, the check is a bounded operational criterion
under a faithful SpecIR encoding, not a theorem of program safety.


%=============================================================================
\section{Summary}
\label{sec:conc:summary}
%=============================================================================

\paragraph{Problem.}
Thin specification regions---especially catch-all residuals and ordered
overlaps---escape everyday unit suites; unstructured feedback does not name
the broken guard.

\paragraph{Evidence.}
Conjunctive Accept cuts escaped defects at comparable mean conformance.
Prevention evidence is industrial-first: Screen cuts FAR from 56.8\% to
32.6\% on $n{=}132$ mutants; the pooled 8.8\% to 5.0\% reading is aggregate
only and is not the release claim.
When repair still has headroom, typed feedback rescues the failures that
test-only dumps miss (E6: on test-only Conf.${<}1$, ${+}8.2$\,pp; all-task
supporting aggregate ${+}7.7$\,pp; semantic\_ir $86.9\%$ vs.\ test\_only
79.1\%; W/L/T $14/4/102$; 95\% CI $[2.3,\,13.6]$\,pp; $p{=}0.018$;
all 14 Conf.\ wins are \texttt{others} rescues, none are ordering repairs;
14/18 decisive wins, 77.8\%; Wilson 95\% CI $[54.8,\,91.0]$).
When test-only Conf.\ is already partial ($0{<}\mathrm{Conf}{<}1$), keep B2
rather than forcing Semantic IR (${-}2.5$\,pp).
SIFR is not claimed superior to every structured feedback channel (E14:
\texttt{execution\_trace\_matched} 85.4\% vs.\ \texttt{semantic\_ir} 75.1\%).
The controlled causal estimate for SIFR remains E6 on BENCH-120; Gemini and
filtered external hard seeds are protocol-transfer diagnostics, not a second
${+}7.7$\,pp headline.
Equal-$K$ Conf.\ ranking places \texttt{M\_eq} at $100.0\%$ vs.\ B2 $97.5\%$
(${+}2.5$\,pp; $3/0/117$) as a matched-budget check, not a substitute for
E6's within-M feedback isolation.
The public ACL case shows the operational signature: tests green, Accept red.

\paragraph{Deployment.}
Prefer test-feedback (B2) by default; escalate to full mode~M when release
requires $\mathrm{Accept}{=}1$ or a contractual FAR budget exists \emph{and}
headroom remains (Table~\ref{tab:deployment-boundary}).
Screen helps on industrial-pattern mutants but does not attain FAR${\le}5\%$
on that slice.
When B2 already achieves Accept${=}1$ (GPT-4o/Claude industrial-pattern:
B2 100.0\% vs.\ M 95.5\%), do not escalate.
When contracts are simple enough that B2 saturates, the cheaper loop is enough.
The pre-hoc HTE checklist (precision 0.167) is a deployment policy, not a
fourth algorithmic contribution.

\paragraph{Scope.}
Gains are headroom-dependent.
Near-ceiling Conf.\ rankings belong in the appendix as secondary stress
material, not as the safety headline.


%=============================================================================
\section{Key Findings}
\label{sec:conc:takeaways}
\label{sec:conc:remember}% retained alias for older cross-refs
%=============================================================================

\begin{enumerate}
  \item Score release by FAR/PDR under conjunctive Accept
        ($\mathrm{Conf}{=}1$ and $\mathrm{Screen}{=}\mathit{pass}$), not by a
        higher mean pass rate alone; read E2 industrial FAR first, then pooled
        aggregate.
  \item Typed feedback matters as catch-all rescue under collapsed test-only
        Conf.: the E6 win profile on BENCH-120 is \texttt{others} residual
        rescue (14/14; 14/18 decisive); the partial-failure band does not
        favour typed IR (${-}2.5$\,pp); keep B2 there rather than forcing
        Semantic IR; ordering defects motivate $\Phi_i$ witnesses and E2,
        not the E6 Conf.\ delta.
  \item Pay for the full loop only when the release bar and headroom demand
        it (Table~\ref{tab:deployment-boundary}); escalate on Accept or a
        contractual FAR budget, not on pooled FAR${\le}5\%$ attainment;
        B6 prevention equals B2/A2 by design when Screen is off (HardSynthetic
        FAR 0\%; industrial B6 unavailable in \texttt{prevention\_full\_v1}).
\end{enumerate}


%=============================================================================
\section{Future Research Directions}
\label{sec:conc:future}
%=============================================================================

\paragraph{Richer encodings.}
Extend first-match witness generation beyond the present LIA fragment to
temporal properties, stateful accumulation, and nonlinear arithmetic---the
residues that single-scenario $\Phi_i$ suites still miss.

\paragraph{Stronger Screen instances.}
Replace the PoC smell pack with CI analysers or bounded proof certificates on
candidates that already clear Conf., keeping the Accept conjunction fixed.

\paragraph{Adaptive budgets.}
Early-stop easy tasks; spend repair iterations where test-only Conf.\ collapses,
so escalation to~M tracks the release bar rather than a fixed $K$.

\paragraph{Live industrial replication.}
Instrument practitioner SOFL/FSF projects against their own acceptance
criteria, beyond the reconstructed public ACL, harvest, and desensitized
pilots already reported.


%=============================================================================
\section{Closing Remarks}
\label{sec:conc:closing}
%=============================================================================

\textsc{SgDP} is not a substitute for full formal verification.
It is a practical decision framework: structured feedback and conjunctive
release keep many specification-conformance defects out of the ship gate on
ordered-guard workflows, while remaining usable inside a sprint.
The evidence that matters is E2 (industrial FAR 56.8\%$\to$32.6\% primary;
pooled PDR/FAR aggregate only), E6 on BENCH-120 (headroom strata and
catch-all rescue), equal-$K$ Conf.\ ranking, the ACL interception case, and
Table~\ref{tab:deployment-boundary}.
Filtered external hard seeds support protocol transfer under test-only
Conf.${<}1$; they do not replace the controlled E6 causal estimate.

